---
title: 'Exercise: R Markdown'
author: 'Veronika Batzdorfer'
date: '2025-06-23'
---

```{r child = "./content/config/exercises_setup.Rmd"}
```

The structure of this final exercise for the course is a bit different. Create and work with `R Markdown` documents (creating `HTML` output) to go through some of the things we covered and did in the previous sessions. There are no coding tasks in this document.

We have created an `R Markdown` document (which we have also already knitted to create an `HTML` output file) that demonstrates some of the things you can do/create with `R Markdown` and repeats a few of the topics and steps we went through in the sessions before this one.

This document uses *Gapminder* data and you can find it in the `exercises` folder: It is called `explore_gapminder.Rmd`.

```{block, box.title = "1", box.body = list(fill = "white"), box.icon = "fa-star"}
The first thing we want you to do is to open the `explore_gapminder.Rmd`file in *RStudio* and explore it a bit to see what it contains. You can also open `explore_gapminder.html` (in your browser) to see the `.Rmd` and the resulting output document side-by-side.
```



```{block, opts.label = "clues"}
You can open the `.Rmd` file via the `File` tab or the menu (`File` -> `Open File`) in *RStudio*.
```

You might notice that there are quite a few things specified in the `YAML` header. Let's briefly go through them:

toc: true -> The document will contain a table of contents (ToC)

toc_depth: 3 -> The ToC will contain header levels 1 to 3

number_sections: true -> The sections divided by headers will be numbered

toc_float: true -> The ToC is floating, meaning that it moves when you scroll

code_folding: hide -> By default, the code chunks are hidden, but you display them by clicking the `Code` buttons in the `HTML` document

theme: flatly -> The [*Bootswatch*](https://bootswatch.com/) them [*flatly*](https://bootswatch.com/flatly/) is used to style the document

highlight: tango -> The document uses the [*Pandoc* code highlighting style](https://www.garrickadenbuie.com/blog/pandoc-syntax-highlighting-examples/) *tango*

code_download: true -> The document includes a button allowing you to download the full code

df_print: paged -> When data frames are printed in the document they are printed in paged tables

**NB**: To knit the document you need to have the packages it uses installed. These are the following ones:
`rmarkdown`, `knitr`, `tidyverse`, `visdat`, `janitor`, `pander`, `patchwork`, `correlation`, `GGally`, `broom`, `sjPlot`, `scales`.

An easy option for checking whether you have these packages installed, doing so if that is not the case in one go, and loading them is the `packages()` function from the `easypackages` package. To use it for this purpose, you can run the following code:

```{r easyp, eval = F}
if (!require(easypackages)) install.packages("easypackages")
library(easypackages)

packages("rmarkdown", "knitr", "tidyverse", "visdat", "janitor", "pander", "patchwork", "correlation", "GGally", "broom", "sjPlot", "scales", prompt = F)
```

Feel free to play around a bit with the `explore_gapminder.Rmd` and its output (you can also change parts of the `YAML` header to see how that influences the output).

```{rblock, box.title = "2", box.body = list(fill = "white"), box.icon = "fa-star"}

Use data transformations and visualizations to answer the following questions.

a) Which five countries had the largest population size in 2018 and how has the population size changed since 1960 ?
  
```

```{r two-exercise, solution = TRUE}

top5_countries <- gap_full %>%
     filter(year == 2018) %>% 
     arrange(-pop) %>% 
     select(country) %>% 
     head(5) %>% 
     pull()

gap_full %>% 
  filter(country %in% top5_countries) %>% 
  ggplot() +
  geom_line(mapping = aes(x = year, y = pop, color = country))+
  theme_bw()

```


```{rblock, box.title = "3", box.body = list(fill = "white"), box.icon = "fa-star"}


b) Rank fertility rate (`fert`) in 2018 for countries like Vietnam, Poland, Afghanistan, Armenia, Denmark
c) What is the relationship between GDP and fertility ? Visualize it.
  
```


```{r three-exercise, solution = TRUE}

gap_full %>%
  filter(year==2018, country %in% c("Poland", "Afghanistan", "Armenia", "Vietnam", "Denmark")) %>%
  arrange(fert) %>%
  select(country, fert) %>% 
  knitr::kable()


gap_full %>%
  filter(year==2018) %>%
  ggplot(aes(y=fert, x=gdp_percap)) +
  geom_point() +
  geom_smooth(method = "lm") +
  scale_x_continuous(limits = c(0, 100000)) +
  theme_bw()

```


```{rblock, box.title = "4", box.body = list(fill = "white"), box.icon = "fa-star"}


d) If you look at the continent level relationship between per-capita GDP and life_expectancy (`life_exp`) what are the trends for continents in 2018 ? Visualize it.
  
```



```{r four-exercise, solution = TRUE}

gap_full %>%
  filter(year==2018) %>%
  drop_na(life_exp, gdp_percap, continent) %>%
  ggplot(aes(y=life_exp, x=gdp_percap, color=continent)) +
  geom_point() +
  geom_smooth(se=F, method = "lm") +
  scale_x_log10() +
  facet_wrap(~continent, scales = "free_y") +
  theme_bw()

```




```{block, box.title = "5", box.body = list(fill = "white"), box.icon = "fa-star"}
The big task for this exercise is to use the workflow with the `titanic` data set. (see in data/titanic.csv, and the codebookd titanic_codebook.csv). Using the `explore_gapminder.Rmd` as a starting point and guidance, do the following in this document:

1. Load and wrangle the data the same way as before to create a subset in the exploratory data analysis (e.g., select variables like demographic variables, handle missing values, recode values, and calculate aggregate measures).

Select key demographic and survival-related variables: `Survived`, `Pclass`, `Sex`, `Age`, `Fare`, `Embarked`, `SibSp`, `Parch`

2. Get an overview of the missing data using the `vis_miss()` function from the `visdat` package.

4. Explore variable distributions. Look at the relative frequencies for the variables using a function from the `janitor` package for `Sex`, `Survived`, `Pclass`.

5. Create bar plots with `ggplot2` to visualize the relative frequencies (percentages) for the variables `Pclass` and `Sex`.

6. Using the `pander` package, include a table with the output of the `base R` function `summary()` for the variables like `Sex`, `Fare`.


7. Create a `ggplot2` boxplot to show differences differences in Age by Survived, and differences in Fare by Pclass and Sex.

8. Calculate correlations for continous numeric variables (`Age`, `Fare`)

9. Create a plot with the `GGally` package to visualize these correlations.

10. Calculate a logistic regression model using `glm()` with `Survived` as the outcome variable. Include predictors like: Pclass, Sex, Age, Fare, SibSp, Parch

11. Visualize the regression results using the sjPlot package.



This is a lot, but you can find template code (`explore_gapminder.Rmd`) for most of this in the the `solutions` folder (and the rest in the slides for the previous sessions). 
```


