---
title: " Data Literacy: Introduction to R "
subtitle: "Exploratory Data Analysis"
author: "Veronika Batzdorfer"
date: "2026-05-08"
presenter: Veronika
editor_options: 
  chunk_output_type: console
---
layout: true 

```{r child = "content/config/sessions_setup.Rmd"}
```



---

## Exploring EDA

In this session, we will focus on the following:

- Compute summary statistics for numeric variables

- Produce counts, proportions and crosstabs for categorical variables

- Use group_by() + summarize() and across() for group summaries

- Run simple correlation analyses and interpret them

- Create concise EDA pipelines and diagnostic plots


---

## Disclaimer: A flood `r ji("ocean")` of packages `r ji("package")`

In the previous sessions, we have tried to focus on a small set of packages (mostly from `base R` and the `tidyverse`) and took a deep dive `r ji("goggles")` exploring their functionalities. By comparison, in this session, we will browse through more different packages. 



---

## Wrangling pipeline:  Load packages & prepare data

.small[
```{r load-data, message = F, warning=FALSE}

library(tidyverse)   # dplyr, ggplot2, tidyr, etc.
library(janitor)     # tabyl, adorn_*
library(corrr)       # neat correlation tibbles
library(broom)       # tidy test results 

# prepare mtcars: convert some numeric columns to factors and relabel
data("mtcars")
df <- mtcars %>%
  rownames_to_column(var = "model") %>%
  mutate(
    cyl = factor(cyl, levels = c(4,6,8), labels = c("4cyl","6cyl","8cyl")),
    gear = factor(gear),
    am   = factor(am, levels = c(0,1), labels = c("automatic","manual"))
  ) %>%
  as_tibble()

glimpse(df)
```

```{r select-rename}
df %>%
  summarize(
    mean_mpg = mean(mpg),
    sd_mpg   = sd(mpg),
    median_hp = median(hp),
    iqr_wt    = IQR(wt),
    min_qsec  = min(qsec),
    max_qsec  = max(qsec)
  ) %>%
  round(2)

```
]

---
## `dplyr::group_by()`

The `dplyr` function `group_by()` creates data frames (tibbles) that are grouped by one or more variables. 

```{r group-by, eval=FALSE}

df %>%
  group_by(cyl) %>%
  summarize(
    n = n(),
    mean_mpg = mean(mpg),
    sd_mpg   = sd(mpg),
    mean_hp  = mean(hp)
  ) %>%
  ungroup()

```

.right[`r emo::ji("left_arrow_curving_right")`]

---

class: middle

```{r ref.label = "group-by", echo = F}

```

---

## Grouping and ungrouping

Keep in mind that `group_by()` changes the class of a dataframe. To avoid (accidentally) continuing to perform operations groupwise, it is necessary to use the .highlight[`ungroup()`] function (typically at the end of a pipe). 

---

## `dplyr::across()`

Multiple summaries for many columns.  

```{r across, eval=FALSE}
df %>%
  select(cyl, mpg, hp, wt, qsec) %>%
  group_by(cyl) %>%
  summarize(
    across(mpg:qsec,
           list(mean = ~mean(.x), sd = ~sd(.x)),
           .names = "{col}_{fn}")
  ) %>%
  ungroup()

```

.right[`r emo::ji("left_arrow_curving_right")`]

---

class: middle

```{r ref.label = "across", echo = F}

```

---

## Frequencies: `table()`

A simple way of looking at frequencies (e.g., for categorical variables) that we have already seen before is the `base R` function `table()`.

```{r table}
table(df$cyl)
```

If you also want to include `NA` in the frequency counts, you need to specify the argument `useNA = "always"`.

```{r table-NA}
table(df$cyl, useNA = "always")
```

---

## Proportions with `tidyverse`


```{r prop-table}

# tidyverse approach
df %>%
  count(cyl) %>%
  mutate(prop = n / sum(n))
```

---


## Frequencies and proportions with `janitor::tabyl()`

The [`janitor` package](https://github.com/sfirke/janitor) provides the [`tabyl()` function](https://sfirke.github.io/janitor/articles/tabyls.html) for creating frequency and proportion tables:


```{r tabyl, message = F}
library(janitor)

df %>%
  tabyl(cyl) %>%
  adorn_pct_formatting(digits = 1)
```

---
## Tables in `R`

There are typically two types of outputs you can produce with `R` for further use in reports or publications: tables and plots. 

However, as summary statistics, frequencies, and proportions are often presented in tables, we will briefly discuss how to create tables as output in `R` in the following.

As with almost everything in `R`, there are many options (read: packages) for creating tables. We will show examples from three popular options:
- [`stargazer`](https://cran.r-project.org/web/packages/stargazer/index.html)
- [`gtsummary`](http://www.danieldsjoberg.com/gtsummary/) + [`flextable`](https://davidgohel.github.io/flextable/index.html)

---

## Summary tables with `stargazer`

If we, e.g., want to create a summary statistics table as text output (e.g., for printing in the `R` console), we can use `stargazer` for that. 

**NB**: As the main `stargazer()` function does not work with tibbles, we need to convert our data to a regular data frame. 

```{r stargazer-text, eval=FALSE}
library(stargazer)

df  %>% 
  select(mpg,
         cyl,
         gear) %>% 
  as.data.frame() %>% 
  stargazer(type = "text",
            digits = 2,
            title="Descriptive statistics")
```

.right[`r emo::ji("left_arrow_curving_right")`]

---

class: middle

```{r ref.label = "stargazer-text", echo = F}

```


---

## Relationships between variables

In the following, we will briefly discuss some options for two methods of exploring relationships between variables:

- crosstabulation (for categorical variables)

- correlations (for numeric and/or binary variables)

---

## Crosstabs

You can use the `base R` functions `table()` and `prop.table()`

```{r base-crosstabs}

# base R contingency table
tab <- table(df$cyl, df$am)
tab

# row percentages
round(prop.table(tab, margin = 1) * 100, 1)

```



---

## Crosstabs with the `janitor` package

The `tabyl()` function from the `janitor` package provides quite a few options for crosstabs. 

.small[
```{r tabyl-crosstabs}
library(janitor)

df %>%
  tabyl(cyl, am) %>%
  adorn_totals("both") %>%
  adorn_percentages("row") %>%
  adorn_pct_formatting(digits = 1) %>%
  adorn_ns(position = "front")


```
]

---

## Other options for crosstabulation in `R`

As with most things in `R`, there are many options for creating crosstabs. Some alternatives to the ones we presented before include the `CrossTable()` and `crosstab()` functions from the [`descr` package](https://cran.r-project.org/web/packages/descr/index.html) or the `ctable()` function from the [`summarytools` package](https://github.com/dcomtois/summarytools).

Another quite very versatile option for creating crosstabs as well as output based on those (e.g., in combination with the `flextable` package we have briefly used before) is the [`crosstable` package](https://danchaltiel.github.io/crosstable/)

---

## Chi-Square for independence (categorical)

We can use the `summary()` function from `base R` in combination with `table()` to perform a chi-square test.


```{r chi-square-base}
#--Are cyl and am independent? ------

# contingency table for cyl vs am
tbl <- df %>% count(cyl, am) %>% 
  pivot_wider(names_from = am, values_from = n, values_fill = 0)

chisq_res <- janitor::chisq.test(as.matrix(tbl %>% select(-cyl)))

chisq_res


```

---


```{r}
tidy(chisq_res) # broom::tidy for neat display
```



---

## Correlations with `base R`

The `base R` function `cor()` computes the correlation coefficient(s) between two or more variables. This function can be used to calculate *Pearson's r*, *Kendall's tau*, and *Spearman's rho*. We also need to specify how we want to deal with missing values (e.g., use pairwise complete observations). 

.small[
```{r corr-base}
nums <- df %>% select(mpg, hp, wt, qsec, disp)

# base R correlation matrix (pairwise complete)
cor(nums, use = "pairwise.complete.obs", method = "pearson")


```
]


---


```{r}
# corrr style (tidy)
nums %>%
  correlate() %>%
  stretch() %>%            # long form
  arrange(desc(abs(r)))
```


---

## The `corrr` package

The [`corrr` package](https://corrr.tidymodels.org/) is part of the [`tidymodels` suite of packages](https://www.tidymodels.org/) and provides various functions for displaying correlations. The main function is `correlate()` which produces a `tibble` as output.

.small[
```{r corr}
library(corrr)

correlate(nums)
```
]

---

## The `corrr` package

The `corrr` package provides several functions for tweaking/optimizing the output of the `correlate()` function. Here's one example:

.small[
```{r corr-output-tweaks}
nums %>% 
  correlate() %>% 
  shave() %>% 
  fashion()
```
]


---
## Correlation diagnostics

```{r}
# scatter with smoothing
ggplot(df, aes(x = wt, y = mpg, color = cyl)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal()

# pairs plot quick overview
pairs(select(df, mpg, hp, wt, qsec), main = "Pairs for selected vars")

```


---

## Guilty by ~~association~~ correlation

While correlation coefficients are useful for exploring relationships between variables, they can also be misleading. For example, if we do correlation analysis and encounter a (Pearson's) correlation coefficient close to 0, we often think of relationships as pictured below.

```{r dino-plot-1, echo = FALSE, out.width = "45%"}
library(datasauRus)

datasaurus_dozen %>% 
  filter(dataset == "h_lines") %>% 
  ggplot(aes(x = x, y = y)) + 
  geom_point() + 
  theme_classic()

```

---

## Guilty by ~~association~~ correlation

This data set has **the same correlation coefficient (Pearson's r of -0.06)** as the one on the previous slide:

```{r dino-plot-2, out.width = "50%", echo = F}
datasaurus_dozen %>% 
  filter(dataset == "slant_up") %>% 
  ggplot(aes(x = x, y = y)) + 
  geom_point() + 
  theme_classic()
```

---

## Guilty by ~~association~~ correlation `r emo::ji("T-Rex")`

So does this one...

```{r dino-plot-3, out.width = "60%", echo = F}
datasaurus_dozen %>% 
  filter(dataset == "dino") %>% 
  ggplot(aes(x = x, y = y)) + 
  geom_point() + 
  theme_classic()
```

---

## Guilty by ~~association~~ correlation

We could go on... The previous three examples all come from the [`datasauRus` package](https://jumpingrivers.github.io/datasauRus/) which essentially is an extension of [Anscombe's quartet](https://en.wikipedia.org/wiki/Anscombe%27s_quartet) and includes 13 data sets with the same (Pearson) correlation between x and y.

```{r dino-plot-4, out.width = "50%", echo = F}
datasaurus_dozen %>% 
ggplot(aes(x = x, y = y, colour = dataset)) +
  geom_point() +
  theme_void() +
  theme(legend.position = "none") +
  facet_wrap(~dataset, ncol=4)
```

---

## Trust no singular value!

Importantly, the x- and y-variables in these `datasaurus_dozen` data set also all have the same means and standard deviations.

```{r datasaurus-summary-stats, eval=TRUE}
datasaurus_dozen %>% 
  group_by(dataset) %>%
  summarize(
    mean_x = mean(x), 
    mean_y = mean(y), 
    sd_x = sd(x), 
    sd_y = sd(y), 
    corr = cor(x, y, method = "pearson")
  )

```

.right[`r emo::ji("left_arrow_curving_right")`]

---

class: middle

.small[
```{r ref.label = "datasaurus-summary-stats", echo = F}

```
]

---

## Plot your data!

The message from the `datasaurus_dozen` examples should be clear. Relying only on singular values that summarize the location or spread of a single variable or the association of two variables is not a good idea. To avoid reducing a ~~mountain to a molehill~~ dinosaur to a lack of correlation, it is important to plot your data to explore:

- univariate distributions

- grouped univariate distributions (if you want to compare groups)

- bivariate distributions

For that reason (and because it is fun and `R` has a lot to offer in that regard), we will dive `r ji("goggles")` 

---


## Other packages for EDA

Besides the ones we have covered in this session and `summarytools`, there still are plenty of others that also provide some interesting functionalities. Some of these additional/alternative options are:
- [`inspectdf`](https://alastairrushworth.github.io/inspectdf/)
- [`skimr`](https://docs.ropensci.org/skimr/)
- [`descriptr`](https://descriptr.rsquaredacademy.com/index.html)
- [`DataExplorer`](https://boxuancui.github.io/DataExplorer/)
- [`explore`](https://github.com/rolkra/explore)
- [`dataReporter`](https://github.com/ekstroem/dataReporter)

---

## Automated EDA reports

Some of the EDA packages provide functions for generating automated EDA reports, for example:
- the `dfSummary()` function from the `summarytools` package
- the `skim()` function from the [`skimr` package](https://docs.ropensci.org/skimr/)
- the `make_report()` function from the [`dataReporter` package](https://github.com/ekstroem/dataReporter)
- the `report()` function from the [`explore` package](https://github.com/rolkra/explore)

The function `explore`() from the `explore` package also allows for interactive data exploration.

---

# Extracurricular activities

Check out the appendix slides on exploring missing data and outliers and/or watch the [*YouTube* tutorial *EDA with the tidyverse: Exploring Variation*](https://www.youtube.com/watch?v=26hbyVb00xs) by Mauro Lepore for even more EDA inspiration. 
