---
title: "Data Literacy: Introduction to R "
subtitle: "Confirmatory Data Analysis"
author: "Veronika Batzdorfer"
date: "2025-11-21"
presenter: Veronika
editor_options: 
  chunk_output_type: console
---
layout: true 

```{r child = "./content/config/sessions_setup.Rmd"}
```


---

## Content of this session
.pull-left[
**What we will cover**
- Using (more or less) simple regression models

- OLS, GLM, and the like

- How to re-use the results of these models

- How to feed these results into tables
]

.pull-right[
**What we won't cover**
- Theory (and history) of hypothesis testing

- Crazy complex models with elaborated estimators
- e.g., no multilevel models
- also no clustered standard errors

- Bayesian statistics
]

```{r load-data, echo=FALSE}

library(palmerpenguins)

penguins_clean <- penguins %>% drop_na()

```



---

## Last sessions

- We learned how to import/ export data
- We learned how to wrangle data, subsets, create new variables, recode, merge, pivot

---
## Recap Distribution

```{r fig.width= 5, fig.height=4}
#plot first variable
hist(penguins_clean$bill_length_mm, 
     col=rgb(0,0,1,0.25), 
          breaks=10,
     xlab='Bill length [mm] ',
     main='Distribution'
) 

```


---
## Descriptive Statistics
### Select continuous variables for summary
```{r}
library(modelsummary)

df.s <- penguins_clean %>% 
  select(bill_length_mm, body_mass_g, flipper_length_mm) 
 
datasummary_skim(df.s, output = 'flextable')

```

---

## Visualize Bivariate Relationships

```{r fig.width= 6, fig.height=4}
library(ggplot2)

# Scatter pl. with trend line----
ggplot(penguins_clean, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(aes(color = species), alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "black", linetype = "dashed") +
  labs(title = "Body Mass by Flipper Length",
       subtitle = "Clear positive linear relationship",
       x = "Flipper Length (mm)",
       y = "Body Mass (g)")
```

Tip: `facet_wrap()` can show separate relationships by group!


---

## Simple Linear Regression: Our First Model


```{r}
# Estimate the model
penguin_lm_simple <- lm(
  body_mass_g ~ flipper_length_mm,
  data = penguins_clean
)

# View coefficients
penguin_lm_simple
```
---

## Interpreting Model Output: What Does It Mean?

```{r}

summary(penguin_lm_simple)
```


---

## Interpreting Coefficients with `parameters`

```{r}
library(parameters)

model_parameters(penguin_lm_simple)
```


---

## Formula Syntax in `R` 


```{r eval = FALSE}
y ~ x1 + x2 + x3
```

*Note*: Formulas represent a specific object class in `R`.

```{r}
class(y ~ x1 + x2 + x3)
```

---

## Denoting the left-hand side with `~`

In `R`, stating what your dependent variable is very similar to common mathematical notation:

$$y \sim N(\theta, \epsilon)$$

It states that a specific relationship is actually _estimated_, but we, fortunately, don't have to specify errors here.

```{r eval = FALSE}
y ~ x1 + x2 + x3
```

Yet, sometimes it may be a good idea to at least explicitly specify the intercept as here:

```{r eval = FALSE}
y ~ 1 + x1 + x2 + x3
```

---

## Intercept

We can also estimate models without an intercept:

```{r eval = FALSE}
y ~ x1 + x2 + x3 - 1
```

Or intercept-only models as well:

```{r eval = FALSE}
y ~ 1
```

---

## Adding predictors with `+`
You can add as many predictors/covariates as you want with the simple <span style="color:orange">`+`</span>
 operator. See:

```{r eval = FALSE}
y ~ 1 + x1 + x2 + x3 + x4 + x5 + x6 + x7 + x8 + x9 + x10 + x11 + x12 +
  x13 + x14 + x15 + x16 + x17 + x18 + x19 + x20 + x21 + x22 + x23 + x24
```

There's also a shortcut for using all (remaining) variables in a data set as predictors:

```{r eval = FALSE}
y ~ .
```

---

## Multiple Regression: Adding More Predictors

```{r}
penguin_lm_multi <- lm(
  body_mass_g ~ flipper_length_mm +
    bill_length_mm +
    bill_depth_mm +
    species,
  data = penguins_clean
)

# Compare with simple model
anova(penguin_lm_simple, penguin_lm_multi)
```

Why multiple?

---

## Interaction effects with `*` and `:`

We can also easily add interaction effects to a model. As this is the same as multiplying predictor variables, we also use the <span style="color:orange">**`*`**</span> sign for that.

```{r eval = FALSE}
y ~ x1 * x2
```

The code above creates a model formula that includes both the <span style="background-color: yellow">main effects</span> of `x1` and `x2`, and their interaction denoted by <span style="background-color: yellow">`x1:x2`</span>. We can even be more explicit and write that into the formula directly:

```{r eval = FALSE}
y ~ x1 + x2 + x1:x2
```


---

## Interaction Effects with Penguins
Does the effect of flipper length on body mass differ by species?


```{r}
penguin_lm_interact <- lm(
  body_mass_g ~ flipper_length_mm * species,
  data = penguins_clean
)

```
---
## Interaction Effects
```{r}
model_parameters(penguin_lm_interact)
```


**Interpretation**: The flipper_length_mm:speciesChinstrap interaction means the slope is 8.9g/mm less steep for Chinstraps vs Adelie.
Visualize interactions to aid interpretation!

---

## **Transforming** variables within a formula

For simple transformations this can be done as follows:

```{r eval = FALSE}
y ~ log(x)   # computes the log10 for x
y ~ scale(x) # z-transformation of x
```

We could also change the data type of variables within a function, e.g., by converting a numeric variable to a factor using `as.factor(x)`. 

---

## Transforming variables within a formula

If you cannot use a specific function for your tansformation, you have to wrap the operation in the <span style="background-color: yellow">`I()`</span> function. 

```{r eval = FALSE}
y ~ x + I(x^2) # add a quadratic term of x
```

*Note*: Of course, there are also functions in `R` for transforming variables (e.g., standardizing or centering) before we use them in a formula. Besides the `base R` function `scale()` the [`datawizard package`](https://easystats.github.io/datawizard/), e.g., provides a few functions for that.

---

## Where to use formulas?

The previous descriptions mainly refer to formulas used in regression models in `R`. However, formulas are also used in other hypothesis testing methods that distinguish between dependent and variables, such as t-tests or ANOVA. We will try out some of those in the following...

---

## Testing group differences in the distribution
### T-Test: Two Groups

A common method for analyzing group differences are t-tests. You can use the two-sample <span style="background-color: yellow">`t.test()`</span> function:

```{r t-test}
#  two-sample t-test
# Research question: Do male and female penguins differ in body mass?

t_test_sex <- t.test(
  body_mass_g ~ sex,          # Outcome ~ Grouping variable
  data = penguins_clean,      
  alternative = "two.sided"   # Two-tailed test 
)

```

---
## T-Test Results

```{r}
print(t_test_sex)

```

.small[
*Note*: By default, `R` uses [Welch's t-test](https://en.wikipedia.org/wiki/Welch%27s_t-test) (instead of [Student's t-test](https://en.wikipedia.org/wiki/Student%27s_t-test)) which does not assume homogeneity of variance across groups.
]

---
## Visualize T-Test

.pull-left[

```{r}
t_test <- ggplot(penguins_clean,
                 aes(x = sex, y = body_mass_g,
                     fill = sex)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) +
  geom_jitter(width = 0.2, alpha = 0.5, show.legend = FALSE) +
  labs(
    title = "Body Mass by Sex: Significant Difference",
    subtitle = "Males are ~860g heavier than females (p < 0.001)",
    x = "Sex",
    y = "Body Mass (g)"
  ) +
  scale_fill_manual(values = c(female = "darkorange", male = "steelblue"))
```
]

.pull-right[
```{r}
t_test
```
]

---

## Testing multiple groups: ANOVA

Research question: Do species differ in flipper length?

```{r anova}

peng_clean <- penguins %>% 
  filter(!is.na(flipper_length_mm))

# 1-way ANOVA
fit <- aov(flipper_length_mm ~ species, data = peng_clean)
summary(fit)

```

---

## Anova visualization
.pull-left[
```{r}
anova_plot <- ggplot(peng_clean,
                     aes(x = species,
                         y = flipper_length_mm,
                         fill = species)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Flipper Length by Species",
       y = "Flipper Length (mm)",
       x = "Species") +
  theme_minimal() +
  theme(legend.position = "none")
```

]

.pull-right[
```{r}
anova_plot
```
]


---

## Two-way ANOVA

Species + Sex effects on body mass

```{r anova-summary}

peng_clean <- penguins %>% 
  filter(!is.na(body_mass_g), !is.na(sex),
         !is.na(species))

fit2 <- aov(body_mass_g ~ species * sex,
            data = peng_clean)

summary(fit2)

```

---

## Post-hoc test (Tukey HSD)

To find out which groups differ from each other we can compute *Tukey Honest Significant Differences*.

```{r tukey}

TukeyHSD(fit)
```

---

## Assumptions Check: Residual Diagnostics

.pull-left[
```{r eval=FALSE}
library(broom)

diag_data <- augment(penguin_lm_multi)

p_resid <- ggplot(diag_data, aes(.fitted, .resid)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title = "Residuals vs Fitted",
    x = "Fitted values",
    y = "Residuals"
  ) +
  theme_minimal()

p_qq <- ggplot(diag_data, aes(sample = .std.resid)) +
  stat_qq(alpha = 0.6) +
  stat_qq_line() +
  labs(
    title = "Normal Q-Q Plot",
    x = "Theoretical Quantiles",
    y = "Standardized Residuals"
  ) +
  theme_minimal()

```
]

.pull-right[
```{r , echo=FALSE}
library(patchwork)

(p_resid / p_qq) + plot_layout(heights = c(3, 3))


```
]


---

## Multicollinearity: VIF Check

When predictors correlate highly, coefficient estimates become unstable. Check Variance Inflation Factor (VIF):

```{r}
library(car)
vif(penguin_lm_multi)
```

**Rule of thumb**: VIF > 5 indicates problematic multicollinearity. Here, bill length and depth are somewhat correlated but acceptable.

**Solution**: Remove one predictor, combine them, or use regularization.
---

## Non-paprametric Alternative

What if our data are <span style="background-color: orange">*not normally distributed*</span>? Check **normality** with the: [Shapiro-Wilk](https://en.wikipedia.org/wiki/Shapiro%E2%80%93Wilk_test) test.

.pull-left[
```{r swt}
# Shapiro-Wilk test on residuals
shapiro.test(residuals(penguin_lm_multi))

```
]
.pull-right[
```{r fig.width=4, fig.height=3}
# Histogram of residuals
hist(residuals(penguin_lm_multi), 
     main = "Distribution of Residuals",
     xlab = "Residuals")

```
]
.small[
**Note**: With large samples, even minor deviations become "significant".
Focus on visual diagnostics and practical significance.
]
---

## Non-parametric alternative to the T-test

If the data are not normally distributed, the [Wilcoxon/Mann-Whitney](https://en.wikipedia.org/wiki/Mann%E2%80%93Whitney_U_test). Use when: data is skewed, non-normal, or sample size is small.

.pull-left[
```{r wilcox}
# Do 2 groups differ?
# E.g.: flipper length between 2 species---->

# Filter only 2 species for a valid two-group  test
peng_two <- penguins %>% 
  filter(species %in% c("Adelie", "Gentoo"))

# Wilcoxon rank-sum test 
wilcox.test(flipper_length_mm ~ species,
            data = peng_two)

```
]
.pull-right[
```{r}
#pairwise comparison between all species

pairwise.wilcox.test(
  x = penguins$flipper_length_mm,
  g = penguins$species,
  p.adjust.method = "BH"
)
```
]

---

## Checking regression assumptions

As in the case of ANOVA, check regression assumptions. There are plenty of good tutorials online (such as [this blog post by Ian Ruginski](https://www.ianruginski.com/post/regressionassumptions/) or [this chapter](https://bookdown.org/jimr1603/Intermediate_R_-_R_for_Survey_Analysis/testing-regression-assumptions.html) in the online book [*R for Survey Analysis*](https://bookdown.org/jimr1603/Intermediate_R_-_R_for_Survey_Analysis/)).


---

## Making Predictions 
.small[Models are useful for prediction! Let's predict body mass for a new penguin:]

```{r}
# New observation
new_penguin <- data.frame(
  flipper_length_mm = 200,
  bill_length_mm = 45,
  bill_depth_mm = 15,
  species = "Gentoo"
)

# Point prediction
predict(penguin_lm_multi, newdata = new_penguin)

# With confidence interval (uncertainty in estimate)
predict(penguin_lm_multi, newdata = new_penguin, 
        interval = "confidence")

```

---

## Standardized Effect Sizes (Beta Coefficients)
To compare effect magnitudes across variables with different scales, use standardized coefficients:


```{r}
library(lm.beta)

# Standardize coefficients
penguin_lm_beta <- lm.beta(penguin_lm_multi)

# Extract standardized coefficients
coef(penguin_lm_beta)["flipper_length_mm"]
coef(penguin_lm_beta)["bill_length_mm"]
```

.small[**Note**: A 1 SD increase in flipper length →  0.64 SD increase in body mass – the strongest predictor!]

---
## Categorical Predictors

`R` automatically handles categorical variables (factors) by creating dummy variables. 

```{r}
contrasts(penguins_clean$species)

#model matrix
head(model.matrix(penguin_lm_multi))
```

.small[
- Reference level
- Other levels
]

---
## Generalized linear models
There are a multitude of other estimators/link functions, so-called **family objects**, e.g., binomial logistic or Poisson regression through the `glm()` function. See `?family` for an overview.


- <span style="background-color: orange">logistic</span> (`glm ( y ~ x, family = 'binomial`)) for binary data,
- <span style="background-color: orange">Poisson</span> (`glm ( y ~ x, family = 'poisson`)) for count data,
- <span style="background-color: orange">negative binomial</span> for overdispersed count data (`glm( y ~ x, family = 'quasipoisson`))


---

## GLM with Penguins: Logistic Regression Example

.small[Let's predict species (binary: Adelie vs. not) from measurements.]

```{r log-reg}
# Binary outcome
penguins_binary <- penguins_clean %>%
  mutate(is_adelie = ifelse(species == "Adelie", 1, 0))

# Logistic regression
penguin_glm <- glm(
  is_adelie ~ flipper_length_mm + bill_length_mm + bill_depth_mm,
  family = binomial(),
  data = penguins_binary
)

# Odds ratios
exp(coef(penguin_glm))

```


---

## Assessing model quality

The [`performance` package](https://easystats.github.io/performance/) that is that is part of the [`easystats` collection of packages](https://easystats.github.io/easystats/) offers some functions for assessing model quality, including different types of R². A commonly used metric for logistic regression models, e.g., is Nagelkerke's R².

---

## Model Comparison with AIC

Compare nested models using Akaike Information Criterion – balances fit and complexity.

```{r probit-reg}
# Simple model
model1 <- lm(body_mass_g ~ flipper_length_mm, 
             data = penguins_clean)

# Complex model
model2 <- lm(body_mass_g ~ flipper_length_mm + bill_length_mm + species, 
             data = penguins_clean)

# Compare
AIC(model1, model2)


```

.small[
Lower AIC = better model. The complex model is preferred here.]


---
## Stepwise Selection

Start with a simple model and stepwise remove possible predictors

```{r stepwise, eval=FALSE}
# Stepwise AIC based selection

step <- step(model1, scope = ~ flipper_length_mm + bill_length_mm + species, trace = FALSE)

summary(step)
```

.right[`r emo::ji("left_arrow_curving_right")`]

---

class:middle
.small[
```{r ref.label = "stepwise", echo = FALSE}
```
]


---

## Comparing model performance

The `performance` package we have previously used for calculating Nagelkerke's R² for our logistic regression model also provides a function for comparing model performance.


---

## Other regression variants

While `glm()` already provides quite a few estimators and link functions, depending on the distribution of your dependent variable, you may need more specialized regression models. A few interesting options for that, e.g, include are the [`MASS` package](https://cran.r-project.org/web/packages/MASS/index.html) for negative binomial regression, the [`pscl` package](https://cran.r-project.org/web/packages/pscl/index.html) for zero-inflated (negative) binomial regression and hurdle models, or the [`mcp` package](https://lindeloev.github.io/mcp/) for regression with multiple change points.


---

## Handling regression results

While it is (still) common practice to run regressions, search for 'significant' `p`-values, and paste the results into a table without interpreting them substantially, this may not be the best thing to do.

---

## Reporting Results

As we have seen in the session on *Exploratory Data Analysis*, there are different options for creating tables as output in `R`:  [`stargazer`](https://cran.r-project.org/web/packages/stargazer/index.html) and [`gtsummary`](http://www.danieldsjoberg.com/gtsummary/)

---

## Example: Regression tables with the `stargazer` package

```{r stargazer, eval = FALSE}
library(stargazer)
stargazer(
  penguin_lm_multi, 
  type = "text",        # For console output; use "html" or "latex" for other formats
  title = "Regression Results",
  out.header = FALSE,    
  digits = 3             

```

.right[`r emo::ji("left_arrow_curving_right")`]

---

class: center, middle
.small[
```{r ref.label = "stargazer", echo = FALSE}
```
]

---


## Coefficient Plots (Odds Ratios)

.small[Create coefficient plot]

.pull-left[
```{r coeffic, fig.width= 5, fig.height=4, eval=FALSE}

library(sjPlot)

# Plot odds ratios with 95% CI
plot_model(
  penguin_glm,
  show.values = TRUE,
  value.offset = .3,
  transform = "exp",     # transforms log-odds → odds ratios
  title = "Odds Ratios for Predicting Adelie Species"
)

```
]

.pull-lright[

```{r ref.label = "coeffic", echo = FALSE, fig.width=5, fig.height=6}
```

]

---
## Parameters coefficient plot

```{r fig.width= 5, fig.height=4}

library(parameters)

model_parameters(penguin_glm) %>% 
  plot()+
  theme_minimal() +
  theme(legend.position = "none")

```

---

## Raw Coefficients (Log-Odds)
```{r fig.width=5, fig.height=6}

plot_model(
  penguin_glm,
  show.values = TRUE,
  value.offset = .3,
  transform = NULL,      # keep log-odds
  title = "Log-Odds Coefficients (Logistic Regression)"
)

```


---

## sjPlot: Marginal Effects Plot

```{r fig.width=5, fig.height=6}
plot_model(
  penguin_glm,
  type = "pred",
  terms = "flipper_length_mm",
  title = "Predicted Probability of Adelie vs. Flipper Length"
)

```

---

## You can add multiple predictors

```{r fig.width=5, fig.height=6}
plot_model(
  penguin_glm,
  type = "pred",
  terms = c("flipper_length_mm", "bill_length_mm"),
  title = "Marginal Effects: Interaction Surface"
)

```


---

## Prediction plots vs Regressions coefficients

- Regression coefficients are sometimes a bit hard to read

    -they are based on the scale of the independent variable when unstandardized
    -it is ambiguous what they actually mean substantially
    -for example, is a coefficient of .2 or .05 (practically) meaningful or not?

- Predictions are a solution

    - they basically draw the regression line into the scatter plot of the Y and X variables
    - we can read at which level of X which value of Y is expected

---

## Predictions (sjPlot)

```{r fig.width= 7, fig.height=5}


plot_model(
  penguin_glm,
  type = "pred",
  terms = c("bill_depth_mm","bill_length_mm"),
  title = "Predicted Probability of Adelie by Bill Depth and Length"
)


```

---

## Customized ggplot
.pull-left[
```{r fig.width= 4, fig.height=5}
library(broom)
library(ggplot2)
library(dplyr)

coef_df <- tidy(penguin_glm, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(term = reorder(term, estimate))

coef_p <- ggplot(coef_df, aes(x = term, y = estimate)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high),
                width = .15, linewidth = .8) +
  labs(title = "Logistic Regression Coefficient Plot",
       x = "Predictor",
       y = "Estimate (log-odds)") +
  coord_flip() +
  theme_minimal()


```
]


.pull-right[
```{r echo=FALSE, fig.width=5, fig.height=6}
coef_p
```
]


---

## Summary Table with sjPlot


```{r}
tab_model(
  penguin_glm,
  show.ci = TRUE,
  show.se = TRUE,
  show.aic = TRUE,
  transform = "exp",
  dv.labels = "Predicting Adelie Species"
)

```


---

## Tidy models with `broom`

The [`broom`](https://broom.tidymodels.org/) package provides only 3 but very powerful main functions:
- `tidy()`: creates a `tibble` from your model (coefficients)
- `glance()`: information about your model as `tibble` ('model fit')
- `augment()`: adds observ. information, e.g., individual fitted values to your data

*Note*: With `broom` you can also create a standardized output of other types of models, such as Latent Class Analysis (LCA) (with the package [`poLCA`](https://cran.r-project.org/web/packages/poLCA/index.html)). 

---

## Conflicts

```{r}
# See which packages conflict
conflicts(detail = TRUE)

# Look specifically for tidy conflicts
conflicts(detail = TRUE)$`package:broom`

```

Always load broom after packages like tidyverse to avoid conflicts

---

## `tidy()`

```{r tidy}
# Load in THIS order
library(tidyverse)  
library(palmerpenguins)
library(broom)      
library(parameters)



tidy_penguin <- tidy(penguin_lm_multi)
print(tidy_penguin)

```

---

## `glance()`

```{r glance}

# glance() - model fit
glance(penguin_lm_multi)

```

---

## `augment()`

```{r augment}

# augment() - diagnostics
augment_penguin <- augment(penguin_lm_multi)
head(augment_penguin)

```

---

## From Model to Diagnostics

Combine all three `broom` functions

```{r broom, eval=FALSE}
# 1. Get coefficients (tidy)
coef_table <- tidy(penguin_lm_multi, conf.int = TRUE)

# 2. Get model fit (glance)
fit_stats <- glance(penguin_lm_multi)

# 3. Get diagnostics (augment)
diag_data <- augment(penguin_lm_multi) %>%
  mutate(
    outlier = abs(.std.resid) > 2,
    species = penguins_clean$species  # Add back categorical variable
  )

# 4. Visualize residuals by species
library(ggplot2)
ggplot(diag_data, aes(x = species, y = .resid, fill = outlier)) +
  geom_boxplot() +
  labs(title = "Residuals by Species",
       subtitle = "Outliers highlighted in red",
       x = "Species",
       y = "Residuals") +
  theme_minimal()
```

---
class: center, middle
.small[
```{r}

```{r ref.label = "broom", echo = FALSE}
```

]

---

## Outlook: Other modeling options

As we've said at the beginning of this session, we only looked at some of the basic confirmatory analysis methods. As you can imagine, however, `R` offers plenty of options more more complex statistical models as well. A few examples:

- Structural equation models with [`lavaan`](https://lavaan.ugent.be/)

- Multilevel/mixed-effects models with [`lme4`](https://cran.r-project.org/web/packages/lme4/index.html)

- Bayesian regression models using [`brms`](https://paul-buerkner.github.io/brms/)

- Mediation analysis using [`mediation`](https://cran.r-project.org/web/packages/mediation/vignettes/mediation.pdf)
  - response, mediator, sensitivity analyses (`medsens()`)
---



## Extracurricular activities

As it is such a simple and useful package, we recommend exploring the `broom` package a bit further. Alex Hayes, one authors of the package, [gave an interesting talk about it at the *RStudio* Conference in 2019](https://rstudio.com/resources/rstudioconf-2019/solving-the-model-representation-problem-with-broom/).

---

class: center, middle

# [Exercise](https://rawcdn.githack.com/nika-akin/data-analysis-with-R/983625a7d096079e0598f2d60ca1281c757d4659/exercises/Exercise_4_1_3_Regression_Reporting.html) time `r ji("weight_lifting_woman")``r ji("muscle")``r ji("running_man")``r ji("biking_man")`

## [Solutions](https://rawcdn.githack.com/nika-akin/data-analysis-with-R/983625a7d096079e0598f2d60ca1281c757d4659/solutions/Exercise_4_1_3_Regression_Reporting.html)

























