---
title: "Data Literacy: Introduction to R "
subtitle: "Confirmatory Data Analysis"
author: "Veronika Batzdorfer"
date: "2025-11-21"
presenter: Veronika
editor_options: 
  chunk_output_type: console
---
layout: true 

```{r child = "./content/config/sessions_setup.Rmd"}
```


---

## Content of this session
.pull-left[
**What we will cover**
- Using (more or less) simple regression models

- OLS, GLM, and the like

- How to re-use the results of these models

- How to feed these results into tables
]

.pull-right[
**What we won't cover**
- Theory (and history) of hypothesis testing

- Crazy complex models with elaborated estimators
- e.g., no multilevel models
- also no clustered standard errors

- Bayesian statistics
]

```{r load-data, echo=FALSE}

library(palmerpenguins)

penguins_clean <- penguins %>% drop_na()

```



---

## Last sessions

- We learned how to import/ export data
- We learned how to wrangle data, subsets, create new variables, recode, merge, pivot

---
## Recap Distribution

```{r fig.width= 5, fig.height=4}
#plot first variable
hist(penguins_clean$bill_length_mm, 
     col=rgb(0,0,1,0.25), 
          breaks=10,
     xlab='Bill length [mm] ',
     main='Distribution'
) 

```


---
## Descriptive Statistics

```{r}
library(modelsummary)

df.s <- penguins_clean %>% 
  dplyr::select(bill_length_mm, body_mass_g, flipper_length_mm) 
 
datasummary_skim(df.s, output = 'flextable')

```

---

## Visualize Bivariate Relationships

```{r fig.width= 6, fig.height=4}
library(ggplot2)

# Scatter pl. with trend line----
ggplot(penguins_clean, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(aes(color = species), alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "black", linetype = "dashed") +
  labs(title = "Body Mass by Flipper Length",
       subtitle = "Clear positive linear relationship",
       x = "Flipper Length (mm)",
       y = "Body Mass (g)")
```

facet_wrap() can show separate relationships by species!


---

## Simple Linear Regression: Our First Model


```{r}

# Estimate the model
penguin_lm_simple <- lm(
  body_mass_g ~ flipper_length_mm,
  data = penguins_clean
)

# View basic output
penguin_lm_simple
```
---

## Model Summary: What Does It Mean?

```{r}

summary(penguin_lm_simple)
```


---

## Interpreting Coefficients with `parameters`

```{r}
library(parameters)

model_parameters(penguin_lm_simple)
```


---

## `R` is straightforward and literate

`R` combines the best of two worlds: It is straightforward to write formulas 

```{r eval = FALSE}
y ~ x1 + x2 + x3
```

*Note*: Formulas represent a specific object class in `R`.

```{r}
class(y ~ x1 + x2 + x3)
```

---

## Denoting the left-hand side with `~`

In `R`, stating what your dependent variable is very similar to common mathematical notation:

$$y \sim N(\theta, \epsilon)$$

It states that a specific relationship is actually _estimated_, but we, fortunately, don't have to specify errors here.

```{r eval = FALSE}
y ~ x1 + x2 + x3
```

Yet, sometimes it may be a good idea to at least explicitly specify the intercept as here:

```{r eval = FALSE}
y ~ 1 + x1 + x2 + x3
```

---

## Intercept

We can also estimate models without an intercept:

```{r eval = FALSE}
y ~ x1 + x2 + x3 - 1
```

Or intercept-only models as well:

```{r eval = FALSE}
y ~ 1
```

---

## Adding predictors with `+`
You can add as many predictors/covariates as you want with the simple <span style="color:orange">`+`</span>
 operator. See:

```{r eval = FALSE}
y ~ 1 + x1 + x2 + x3 + x4 + x5 + x6 + x7 + x8 + x9 + x10 + x11 + x12 +
  x13 + x14 + x15 + x16 + x17 + x18 + x19 + x20 + x21 + x22 + x23 + x24
```

There's also a shortcut for using all (remaining) variables in a data set as predictors:

```{r eval = FALSE}
y ~ .
```

---

## Multiple Regression: Adding More Predictors

```{r}
penguin_lm_multi <- lm(
  body_mass_g ~ flipper_length_mm + bill_length_mm + bill_depth_mm + species,
  data = penguins_clean
)

# Compare models
anova(penguin_lm_simple, penguin_lm_multi)
```

Why multiple?

---

## Interaction effects with `*` and `:`

We can also easily add interaction effects to a model. As this is the same as multiplying predictor variables, we also use the <span style="color:orange">**`*`**</span> sign for that.

```{r eval = FALSE}
y ~ x1 * x2
```

The code above creates a model formula that includes both the <span style="background-color: yellow">main effects</span> of `x1` and `x2`, and their interaction denoted by <span style="background-color: yellow">`x1:x2`</span>. We can even be more explicit and write that into the formula directly:

```{r eval = FALSE}
y ~ x1 + x2 + x1:x2
```


---

## Interaction Effects with Penguins
Does the effect of flipper length on body mass differ by species?


```{r}
penguin_lm_interact <- lm(
  body_mass_g ~ flipper_length_mm * species,
  data = penguins_clean
)

model_parameters(penguin_lm_interact)
```

Interpretation: The flipper_length_mm:speciesChinstrap interaction means the slope is 8.9g/mm less steep for Chinstraps than Adelie.
Visualize interactions to aid interpretation!

---

## **Transforming** variables within a formula

For simple transformations this can be done as follows:

```{r eval = FALSE}
y ~ log(x)   # computes the log10 for x
y ~ scale(x) # z-transformation of x
```

We could also change the data type of variables within a function, e.g., by converting a numeric variable to a factor using `as.factor(x)`. 

---

## Transforming variables within a formula

If you cannot use a specific function for your tansformation, you have to wrap the operation in the <span style="background-color: yellow">`I()`</span> function. 

```{r eval = FALSE}
y ~ x + I(x^2) # add a quadratic term of x
```

*Note*: Of course, there are also functions in `R` for transforming variables (e.g., standardizing or centering) before we use them in a formula. Besides the `base R` function `scale()` the [`datawizard package`](https://easystats.github.io/datawizard/), e.g., provides a few functions for that.

---

## Where to use formulas?

The previous descriptions mainly refer to formulas used in regression models in `R`. However, formulas are also used in other hypothesis testing methods that distinguish between dependent and variables, such as t-tests or ANOVA. We will try out some of those in the following...

---

## Testing group differences in the distribution

A common method for analyzing group differences are t-tests. You can use the two-sample <span style="background-color: yellow">`t.test()`</span> function:

```{r t-test}
#  two-sample t-test
# Research question: Do male and female penguins differ in body mass?

t_test_sex <- t.test(
  body_mass_g ~ sex,          # Outcome ~ Grouping variable
  data = penguins_clean,      
  alternative = "two.sided"   # Two-tailed test 
)


print(t_test_sex)

# 4. Extract key statistics 
list(
  t_statistic = round(t_test_sex$statistic, 2),
  df = round(t_test_sex$parameter, 1),
  p_value = t_test_sex$p.value,
  mean_diff = round(t_test_sex$estimate[1] - t_test_sex$estimate[2])
)
```

.small[
*Note*: By default, `R` uses [Welch's t-test](https://en.wikipedia.org/wiki/Welch%27s_t-test) (instead of [Student's t-test](https://en.wikipedia.org/wiki/Student%27s_t-test)) which does not assume homogeneity of variance across groups.
]

---
## Visualize T-Test

```{r}
ggplot(penguins_clean, aes(x = sex, y = body_mass_g, fill = sex)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) +
  geom_jitter(width = 0.2, alpha = 0.5, show.legend = FALSE) +
  labs(
    title = "Body Mass by Sex: Significant Difference",
    subtitle = "Males are ~860g heavier than females (p < 0.001)",
    x = "Sex",
    y = "Body Mass (g)"
  ) +
  scale_fill_manual(values = c(female = "darkorange", male = "steelblue"))
```


---

## Test of normality

What if our data are <span style="background-color: orange">*not normally distributed*</span> in the first place, violating one of the basic assumptions of performing t-tests? To check this: [Shapiro-Wilk](https://en.wikipedia.org/wiki/Shapiro%E2%80%93Wilk_test) test of normality.

```{r swt, eval=FALSE}

# Shapiro-Wilk test on residuals
shapiro.test(residuals(penguin_lm_multi))

# Histogram of residuals
hist(residuals(penguin_lm_multi), 
     main = "Distribution of Residuals",
     xlab = "Residuals")

```

**Note**: With large samples, even minor deviations become "significant". Focus on visual diagnostics and practical significance.

---

## Wilcoxon/Mann-Whitney test

If the data are not normally distributed, the [Wilcoxon/Mann-Whitney](https://en.wikipedia.org/wiki/Mann%E2%80%93Whitney_U_test) test can be a suitable alternative.

```{r wilcox}
# flipper length between 2 species----

# Filter only 2 species for a valid two-group  test
peng_two <- penguins %>% 
  filter(species %in% c("Adelie", "Gentoo"))

# Run the Wilcoxon rank-sum test (non-parametric alternative to t-test)
wilcox.test(flipper_length_mm ~ species, data = peng_two)

```

```{r}
#pairwise comparison between all species

pairwise.wilcox.test(
  x = penguins$flipper_length_mm,
  g = penguins$species,
  p.adjust.method = "BH"
)
```


---

## Testing multiple groups with an ANOVA

There are situations in which we want to test differences across multiple groups. The classic method to use for this is an analysis of variance (ANOVA) and its many variants (ANCOVA, MANOVA, etc.). 

```{r anova}
#One-way anova
# do species differ in flipper length

peng_clean <- penguins %>% 
  filter(!is.na(flipper_length_mm))

# One-way ANOVA
fit <- aov(flipper_length_mm ~ species, data = peng_clean)
summary(fit)

```

---


## Anova visualization

```{r}
ggplot(peng_clean, aes(x = species, y = flipper_length_mm, fill = species)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Flipper Length by Species",
       y = "Flipper Length (mm)",
       x = "Species") +
  theme_minimal() +
  theme(legend.position = "none")
```



---

## Two-way ANOVA: Species + Sex effects on body mass


```{r anova-summary}

peng_clean <- penguins %>% 
  filter(!is.na(body_mass_g), !is.na(sex), !is.na(species))

fit2 <- aov(body_mass_g ~ species * sex, data = peng_clean)

summary(fit2)

```

---

## Post-hoc test (Tukey HSD)

The previous output only indicates that there are statistically significant differences in the groups. To find out which groups differ from each other we can, for example, compute *Tukey Honest Significant Differences*.

```{r tukey, eval=FALSE}

TukeyHSD(fit)
```


---
## Assumption checks
Check assumptions systematically: linearity, normality, homoscedasticity, independence.


```{r eval=FALSE}
library(performance)
library(see)

```


```{r }
par(mfrow = c(1, 2))

# Residuals vs Fitted
plot(penguin_lm_multi, which = 1, 
     main = "Linearity & Homoscedasticity")

# Normal Q-Q plot
plot(penguin_lm_multi, which = 2, 
     main = "Normality of Residuals")

```


---

## Multicollinearity: The VIF Check

When predictors correlate highly, coefficient estimates become unstable. Check Variance Inflation Factor (VIF):

```{r}
library(car)

vif(penguin_lm_multi)
```

**Rule of thumb**: VIF > 5 indicates problematic multicollinearity. Here, bill length and depth are somewhat correlated but acceptable.

**Solution**: Remove one predictor, combine them, or use regularization.

---

## Checking regression assumptions

As in the case of ANOVA, check regression assumptions. There are plenty of good tutorials online (such as [this blog post by Ian Ruginski](https://www.ianruginski.com/post/regressionassumptions/) or [this chapter](https://bookdown.org/jimr1603/Intermediate_R_-_R_for_Survey_Analysis/testing-regression-assumptions.html) in the online book [*R for Survey Analysis*](https://bookdown.org/jimr1603/Intermediate_R_-_R_for_Survey_Analysis/)).


---

## Making Predictions from Your Model
Models are useful for prediction! Let's predict body mass for a new penguin:

```{r}
# New observation
new_penguin <- data.frame(
  flipper_length_mm = 200,
  bill_length_mm = 45,
  bill_depth_mm = 15,
  species = "Gentoo"
)

# Point prediction
predict(penguin_lm_multi, newdata = new_penguin)

# With confidence interval (uncertainty in estimate)
predict(penguin_lm_multi, newdata = new_penguin, 
        interval = "confidence")

# With prediction interval (uncertainty for new observation)
predict(penguin_lm_multi, newdata = new_penguin, 
        interval = "prediction")
```

---

## Standardized Effect Sizes (Beta Coefficients)
To compare effect magnitudes across variables with different scales, use standardized coefficients:


```{r}
library(lm.beta)

# Standardize coefficients
penguin_lm_beta <- lm.beta(penguin_lm_multi)

# Extract standardized coefficients
coef(penguin_lm_beta)["flipper_length_mm"]
coef(penguin_lm_beta)["bill_length_mm"]
```

**Note**: A 1 SD increase in flipper length corresponds to a 0.64 SD increase in body mass – the strongest predictor!

---
## Categorical Predictors: Factor Variables

`R` automatically handles categorical variables (factors) by creating dummy variables. Inspect the design matrix:

```{r}
# See how species is coded
contrasts(penguins_clean$species)

# View model matrix
head(model.matrix(penguin_lm_multi))
```

- Reference level
- Other levels






---
## Generalized linear regressions 
There are a multitude of other estimators/link functions, so-called **family objects**, e.g., binomial logistic or Poisson regression through the `glm()` function. See `?family` for an overview.


- <span style="background-color: orange">logistic</span> (`glm ( y ~ x, family = 'binomial`)) for binary data,
- <span style="background-color: orange">Poisson</span> (`glm ( y ~ x, family = 'poisson`)) for count data,
- <span style="background-color: orange">negative binomial</span> for overdispersed count data (`glm( y ~ x, family = 'quasipoisson`))


---

## GLM with Penguins: Logistic Regression Example

Let's predict species (binary: Adelie vs. not) from measurements – a classification problem.

```{r log-reg}
# Binary outcome
penguins_binary <- penguins_clean %>%
  mutate(is_adelie = ifelse(species == "Adelie", 1, 0))

# Logistic regression
penguin_glm <- glm(
  is_adelie ~ flipper_length_mm + bill_length_mm + bill_depth_mm,
  family = binomial(),
  data = penguins_binary
)

# Odds ratios
exp(coef(penguin_glm))

```

.right[`r emo::ji("left_arrow_curving_right")`]

---

class: middle
.small[
```{r ref.label = "log-reg", echo = FALSE}

```
]

---


## Assessing model quality

The [`performance` package](https://easystats.github.io/performance/) that is that is part of the [`easystats` collection of packages](https://easystats.github.io/easystats/) offers some functions for assessing model quality, including different types of R². A commonly used metric for logistic regression models, e.g., is Nagelkerke's R².


*Note*: The `performance` package also includes several helpful functions for model diagnostics.

---

## Model Comparison with AIC

Compare nested models using Akaike Information Criterion – balances fit and complexity.

```{r probit-reg, eval = FALSE}
# Simple model
model1 <- lm(body_mass_g ~ flipper_length_mm, 
             data = penguins_clean)

# Complex model
model2 <- lm(body_mass_g ~ flipper_length_mm + bill_length_mm + species, 
             data = penguins_clean)

# Compare
AIC(model1, model2)


```

Lower AIC = better model. The complex model is preferred here.
.right[`r emo::ji("left_arrow_curving_right")`]

---

class:middle
.small[
```{r ref.label = "probit-reg", echo = FALSE, eval=TRUE}
```
]

---
## Compare models step-wise using AIC 

Start with a simple model and stepwise remove possible predictors

```{r stepwise}
# Stepwise selection
step_model <- step(model1, scope = ~ flipper_length_mm + 
                     bill_length_mm + bill_depth_mm + species,
                   trace = FALSE)
summary(step_model)

```

.right[`r emo::ji("left_arrow_curving_right")`]

---

class:middle
.small[
```{r ref.label = "stepwise", echo = FALSE}
```
]


---

## Comparing model performance

The `performance` package we have previously used for calculating Nagelkerke's R² for our logistic regression model also provides a handy function for comparing model performance.


---

## Other regression variants

While `glm()` already provides quite a few estimators and link functions, depending on the distribution of your dependent variable, you may need more specialized regression models. A few interesting options for that, e.g, include are the [`MASS` package](https://cran.r-project.org/web/packages/MASS/index.html) for negative binomial regression, the [`pscl` package](https://cran.r-project.org/web/packages/pscl/index.html) for zero-inflated (negative) binomial regression and hurdle models, or the [`mcp` package](https://lindeloev.github.io/mcp/) for regression with multiple change points.


---

## Handling regression results

While it is (still) common practice to run regressions, search for 'significant' `p`-values, and paste the results into a table without interpreting them substantially, this may not be the best thing to do.

---

## Regression tables

As we have seen in the session on *Exploratory Data Analysis*, there are different options for creating tables as output in `R`:  [`stargazer`](https://cran.r-project.org/web/packages/stargazer/index.html) and [`gtsummary`](http://www.danieldsjoberg.com/gtsummary/)

---

## Example: Regression tables with the `stargazer` package

```{r stargazer, eval = FALSE}
library(stargazer)

stargazer(
  penguin_lm_multi, 
  type = "text",        # For console output; use "html" or "latex" for other formats
  title = "Regression Results",
  dep.var.labels = c("AI Trust"),
  covariate.labels = c(
    "Main Branch", 
    "Participation Frequency", 
    "Years Coding"
  ),
  out.header = FALSE,    # Exclude header metadata if outputting to file
  digits = 3             # Number of decimal places
)

```

.right[`r emo::ji("left_arrow_curving_right")`]

---

class: center, middle
.ssmall[
```{r ref.label = "stargazer", echo = FALSE}
```
]


---

## Tidy models with `broom`

The [`broom`](https://broom.tidymodels.org/) package provides only 3 but very powerful main functions:
- `tidy()`: creates a `tibble` from your model (coefficients)
- `glance()`: information about your model as `tibble` ('model fit')
- `augment()`: adds observ. information, e.g., individual fitted values to your data

*Note*: With `broom` you can also create a standardized output of other types of models, such as Latent Class Analysis (LCA) (with the package [`poLCA`](https://cran.r-project.org/web/packages/poLCA/index.html)). 

---

## Conflicts


```{r}
# See which packages conflict
conflicts(detail = TRUE)

# Look specifically for tidy conflicts
conflicts(detail = TRUE)$`package:broom`

```

Always load broom after packages like tidyverse to avoid conflicts

---

## `tidy()`

```{r tidy}
# Load in THIS order
library(tidyverse)  
library(palmerpenguins)
library(broom)      
library(parameters)



tidy_penguin <- tidy(penguin_lm_multi)
print(tidy_penguin)

```

---

## `glance()`

```{r glance}

# glance() - model fit
glance(penguin_lm_multi)

```

---

## `augment()`

```{r augment}

# augment() - diagnostics
augment_penguin <- augment(penguin_lm_multi)
head(augment_penguin)

```

---

## From Model to Diagnostics

Combine all three `broom` functions

```{r}
# 1. Get coefficients (tidy)
coef_table <- tidy(penguin_lm_multi, conf.int = TRUE)

# 2. Get model fit (glance)
fit_stats <- glance(penguin_lm_multi)

# 3. Get diagnostics (augment)
diag_data <- augment(penguin_lm_multi) %>%
  mutate(
    outlier = abs(.std.resid) > 2,
    species = penguins_clean$species  # Add back categorical variable
  )

# 4. Visualize residuals by species
library(ggplot2)
ggplot(diag_data, aes(x = species, y = .resid, fill = outlier)) +
  geom_boxplot() +
  labs(title = "Residuals by Species",
       subtitle = "Outliers highlighted in red",
       x = "Species",
       y = "Residuals") +
  theme_minimal()
```



---

## Outlook: Other modeling options

As we've said at the beginning of this session, we only looked at some of the basic confirmatory analysis methods. As you can imagine, however, `R` offers plenty of options more more complex statistical models as well. A few examples:

- Structural equation models with [`lavaan`](https://lavaan.ugent.be/)

- Multilevel/mixed-effects models with [`lme4`](https://cran.r-project.org/web/packages/lme4/index.html)

- Bayesian regression models using [`brms`](https://paul-buerkner.github.io/brms/)

- Mediation analysis using [`mediation`](https://cran.r-project.org/web/packages/mediation/vignettes/mediation.pdf)
  - response, mediator, sensitivity analyses (`medsens()`)
---



## Extracurricular activities

As it is such a simple and useful package, we recommend exploring the `broom` package a bit further. Alex Hayes, one authors of the package, [gave an interesting talk about it at the *RStudio* Conference in 2019](https://rstudio.com/resources/rstudioconf-2019/solving-the-model-representation-problem-with-broom/).

---

class: center, middle

# [Exercise](https://rawcdn.githack.com/nika-akin/data-analysis-with-R/983625a7d096079e0598f2d60ca1281c757d4659/exercises/Exercise_4_1_3_Regression_Reporting.html) time `r ji("weight_lifting_woman")``r ji("muscle")``r ji("running_man")``r ji("biking_man")`

## [Solutions](https://rawcdn.githack.com/nika-akin/data-analysis-with-R/983625a7d096079e0598f2d60ca1281c757d4659/solutions/Exercise_4_1_3_Regression_Reporting.html)

---
## Coefficient Plots (Odds Ratios)

Create coefficient plot
```{r fig.width= 5, fig.height=4}

library(sjPlot)

# Plot odds ratios with 95% CI
plot_model(
  penguin_glm,
  show.values = TRUE,
  value.offset = .3,
  transform = "exp",     # transforms log-odds → odds ratios
  title = "Odds Ratios for Predicting Adelie Species"
)


```

---
## Parameters coefficient plot

```{r fig.width= 5, fig.height=4}

library(parameters)

model_parameters(penguin_glm) %>% 
  plot()+
  theme_minimal() +
  theme(legend.position = "none")

```

---

## Raw Coefficients (Log-Odds)
```{r}

plot_model(
  penguin_glm,
  show.values = TRUE,
  value.offset = .3,
  transform = NULL,      # keep log-odds
  title = "Log-Odds Coefficients (Logistic Regression)"
)

```


---

## sjPlot: Marginal Effects Plot

```{r}
plot_model(
  penguin_glm,
  type = "pred",
  terms = "flipper_length_mm",
  title = "Predicted Probability of Adelie vs. Flipper Length"
)

```

---

## You can add multiple predictors

```{r}
plot_model(
  penguin_glm,
  type = "pred",
  terms = c("flipper_length_mm", "bill_length_mm"),
  title = "Marginal Effects: Interaction Surface"
)

```


---

## Prediction plots vs Regressions coefficients

- Regression coefficients are sometimes a bit hard to read

    -they are based on the scale of the independent variable when unstandardized
    -it is ambiguous what they actually mean substantially
    -for example, is a coefficient of .2 or .05 (practically) meaningful or not?

- Predictions are a solution

    - they basically draw the regression line into the scatter plot of the Y and X variables
    - we can read at which level of X which value of Y is expected

---

## Predictions (sjPlot)

```{r fig.width= 7, fig.height=5}


plot_model(
  penguin_glm,
  type = "pred",
  terms = c("bill_depth_mm","bill_length_mm"),
  title = "Predicted Probability of Adelie by Bill Depth and Length"
)


```

---

## Customized ggplot

```{r fig.width= 6, fig.height=4}
library(broom)
library(ggplot2)
library(dplyr)

coef_df <- tidy(penguin_glm, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(term = reorder(term, estimate))

ggplot(coef_df, aes(x = term, y = estimate)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high),
                width = .15, linewidth = .8) +
  labs(title = "Logistic Regression Coefficient Plot",
       x = "Predictor",
       y = "Estimate (log-odds)") +
  coord_flip() +
  theme_minimal()


```

---

## Summary Table with sjPlot


```{r}
tab_model(
  penguin_glm,
  show.ci = TRUE,
  show.se = TRUE,
  show.aic = TRUE,
  transform = "exp",
  dv.labels = "Predicting Adelie Species"
)

```


























