---

title: "Data Literacy: Introduction to R"
subtitle: "Excursion Shiny App and LLMs"
author: "otello"
date: "2025-06-24"
presenter: Veronika
output: 
  html_document:
    theme: united
    highlight: tango
runtime: shiny
editor_options: 
  chunk_output_type: console
---


layout: true 

```{r child = "./content/config/sessions_setup.Rmd", warning=FALSE}
```

---

## Shiny apps

> Shiny is an R package that makes it easy to build interactive web apps straight from R. You can host standalone apps on a webpage or embed them in R Markdown documents or build dashboards. You can also extend your Shiny apps with CSS themes, htmlwidgets, and JavaScript actions.

https://shiny.rstudio.com/

---

class: middle

## Example 1: Movie Explorer

.center[https://shiny.rstudio.com/gallery/movie-explorer.html]

---

class: middle

## Example 2: CRAN explorer

.center[https://gallery.shinyapps.io/cran-explorer/]


---

## Why Use Shiny?

- Build interactive applications using only R

- Enables data exploration with dynamic UI

- Deployable as dashboards or embedded in reports

- Popular in health, public sector, finance, academia



---
## Anatomy of a Shiny App


Every Shiny app has 3 key parts:

- .highlight[UI] (user interface)

- .highlight[Server] (logic + reactivity)

- .highlight[Input/output bindings] – Links user actions (**ui**) and app responses (**server**)
  

---

## Let’s explore a simple Shiny interaction 

Pick a number 
```{r}
library(shiny)

# UI element
sliderInput("num", "Pick a number", min = 1, max = 100, value = 50)

```

Result (reactive output)
```{r}
# Output definition
renderText({
  paste("Double of your number is:", input$num * 2)
})

```

---

## Key Concepts in Shiny:

- Reactive .highlight[Inputs]: Values from widgets like sliders, text boxes, select menus

- Reactive .highlight[Outputs]: renderText, renderPlot, renderTable, etc.

- Reactive .highlight[Expressions]: Efficient computations that depend on inputs
---
## Shiny Front-End Anatomy: UI Layout

- `fluidPage()`: most basic layout

  - rows containing columns
  - rows making sure elements appear on the same line

- UI input: Logic: Widget = Web element the user can interact with [Shiny widget gallery](https://shiny.posit.co/r/gallery/widgets/widget-gallery/)


---

### Try this code in your console for a simple UI

```{r}
library(shiny)

# UI layout with clear structure and purpose
ui <- fluidPage(
  titlePanel("Fancy App"),
  
  sidebarLayout(
    
    # Sidebar panel for user inputs or controls
    sidebarPanel(
      h4("Sidebar"),
      p("Use this space for inputs like sliders, selectors, or buttons.")
    ),
    
    # Main panel for outputs and content
    mainPanel(
      h4("Main Content"),
      
      # 3 columns 
      fluidRow(
        column(4,
               h5("Description"),
               p("This area is for explaining what the app does.")
        ),
        
        column(4,
               h5("Model Output"),
               p("Use this section to show summaries, tables, or results.")
        ),
        
        column(4,
               h5("Visualization"),
               p("Render plots or graphics here.")
        )
      )
    )
  )
)

# Server logic placeholder
server <- function(input, output, session) {
  # This is where server-side reactivity will go
}

# Run the app
shinyApp(ui = ui, server = server)

```



---

## Loading and Exploring the Flights Dataset

We’ll use `nycflights13::flights` for a real-world example.

```{r}
library(nycflights13)
library(dplyr)
library(ggplot2)

head(flights)

```

---

## Filter Flights from JFK in January

```{r}
flights %>%
  filter(origin == "JFK", month == 1) %>%
  group_by(carrier) %>%
  summarise(mean_arr_delay = mean(arr_delay, na.rm = TRUE)) %>%
  arrange(desc(mean_arr_delay))

```
---

## TASK 1: Data Wrangling Warm-up (10 min)

Use dplyr to answer:

- Which carrier has the longest average arrival delay?

- What is the shortest average air time from LGA?

---

## Interactive Carrier Selection

Select a carrier below to view data:

```{r}
selectInput("carrier_sel", "Choose a carrier:",
            choices = unique(flights$carrier))

```

Summary Statistics
```{r}
renderPrint({
  req(input$carrier_sel)
  flights %>%
    filter(carrier == input$carrier_sel) %>%
    summarise(
      mean_delay = mean(arr_delay, na.rm = TRUE),
      sd_delay = sd(arr_delay, na.rm = TRUE),
      flights = n()
    )
})

```

---

## Visualize Delays by Carrier

```{r}
renderPlot({
  req(input$carrier_sel)
  flights %>%
    filter(carrier == input$carrier_sel) %>%
    ggplot(aes(x = arr_delay)) +
    geom_histogram(binwidth = 5, fill = "#0073C2FF", color = "white") +
    xlim(-50, 200) +
    labs(title = paste("Arrival Delay for", input$carrier_sel),
         x = "Arrival Delay (min)", y = "Count")
})

```
---

## Add Origin Selection to the Filter

```{r}

selectInput("origin_sel", "Choose origin airport:",
            choices = unique(flights$origin))
```

```{r}
renderPlot({
  req(input$carrier_sel, input$origin_sel)
  flights %>%
    filter(carrier == input$carrier_sel, origin == input$origin_sel) %>%
    ggplot(aes(x = arr_delay)) +
    geom_histogram(binwidth = 5, fill = "#D55E00", color = "white") +
    labs(title = paste("Delays for", input$carrier_sel, "from", input$origin_sel),
         x = "Arrival Delay (min)", y = "Count")
})

```

---

## TASK 2: Multi-Input Filtering (15 min)

Add to the existing app:

- A date filter (hint: use `lubridate::make_date`(year, month, day))

- A scatterplot of `dep_delay` vs. `arr_delay` with color = destination


---

## Let's add a Date Range Input

```{r}
dateRangeInput("date_range", "Select date range:",
               start = "2013-01-01", end = "2013-01-31")

```

```{r}
renderPlot({
  req(input$carrier_sel, input$date_range)
  flights %>%
    filter(carrier == input$carrier_sel) %>%
    mutate(date = as.Date(time_hour)) %>%
    filter(date >= input$date_range[1], date <= input$date_range[2]) %>%
    ggplot(aes(x = dep_delay, y = arr_delay, color = dest)) +
    geom_point(alpha = 0.5) +
    labs(title = "Departure vs. Arrival Delay", x = "Departure Delay", y = "Arrival Delay")
})

```

---

## TASK 3: Build Your Own Mini Dashboard (30 min)

- Choose an aspect of the flights dataset to explore

- Design at least two inputs and two outputs

- Add interactivity (e.g. filtering, comparison, visualization)

---

## Common Mistakes & Debugging Tips

-  Forgetting `req()` before using an input

-  Mixing up `render*` and `output$*`

-  Not wrapping logic inside `reactive()` or render blocks

-  Visuals not appearing? Check for missing `outputId` or input dependencies

---

## Deployment Options

- shinyapps.io — official, free tier available

- Shiny Server (on-premise or cloud)

- Use flexdashboard + runtime: shiny for inline web reports



---
## Resources: Online courses & tutorials

- Overview of resources
  - [*learnR4free*](https://www.learnr4free.com/) by Mine Dogucu
  - the [*Big Book of R*](https://www.bigbookofr.com/) by Oscar Baruffa

- [*swirl* - Learn `R` in `R`](https://swirlstats.com/)

- Learning `R` (and statistics) with a cute story: [Teacups, Giraffes, & Statistics by Hasse Wallum & Desirée de Leon](https://tinystats.github.io/teacups-giraffes-and-statistics/)

---

## Working with text data in `R`

As with almost everything else, there are many great resources for working with text data in `R`. Two good options (and starting points) are:
- the [`tidytext` package](https://juliasilge.github.io/tidytext/) and the "accompanying" book [*Text Mining with R: A Tidy Approach*](https://www.tidytextmining.com/) by Julia Silge & David Robinson
- the [`quanteda` package](https://quanteda.io/) and its accompanying [tutorials](https://tutorials.quanteda.io/)

- a nice and free self-paced online course is [*Text mining in R for the social sciences and digital humanities*](https://tm4ss.github.io/docs/index.html) by Andreas Niekler and Gregor Wiedemann

---

## 'Web development' using `R`

These days, a lot of `R` packages provide tools originally developed for the web. For example:
- [bookdown](https://pkgs.rstudio.com/bookdown/) enables you to publish books written in `R Markdown` online
- [pkgdown](https://pkgdown.r-lib.org/) does the same for your own `R` packages
- [blogdown](https://pkgs.rstudio.com/blogdown/l) is more general and helps you with creating websites (examples to follow)


---

## Writing your own `R` packages

.pull-left[
At a certain point (not now!), you may want to consider writing your own `R` package
- useful for creating reproducible code
- great for distributing your work to others

]

.pull-right[
```{r r-packages, echo = FALSE}
woRkshoptools::include_picture("r_packages.jpg")
```
[Read the book here!](https://r-pkgs.org/)
]

---

class: middle

## It's straightforward in `RStudio`

```{r new-package, echo = FALSE, out.width = "75%"}
woRkshoptools::include_picture("new_package.png")
```

---

## Acknowledgements `r ji("heart")`

All slides were created with the `R` package [`xaringan`](https://github.com/yihui/xaringan) which builds on [`remark.js`](https://remarkjs.com), [`knitr`](http://yihui.name/knitr), and [`R Markdown`](https://rmarkdown.rstudio.com). The exercises were created with the [`unilur` package](https://github.com/koncina/unilur).  
[`woRkshoptools`](https://stefanjuenger.github.io/woRkshoptools/)

