---
title: "Data Literacy: Introduction to R"
subtitle: "Local LLMs & Web Scraping"
author: "Veronika Batzdorfer"
date: "2025-11-21"
presenter: Veronika
output: 
  html_document:
    theme: united
    highlight: tango
runtime: shiny
editor_options: 
  chunk_output_type: console
---


layout: true 

```{r child = "./content/config/sessions_setup.Rmd", warning=FALSE}
```

---

## Objectives


```{r objectives, echo=FALSE}


# Learning Objectives - Keep focused & achievable
objectives <- data.frame(
  Time = c("10min", "10min", "10min", "10min"),
  Activity = c(
    "Setup LM Studio",
    "Query LLMs from R",
    "Scrape & parse Wikipedia",
    "Build hybrid LLM+data pipeline"
  )
)
print(objectives, row.names = FALSE)
```

---

## Prerequisites

```{r}
# Run in R console
required <- c("httr2", "rvest", "dplyr", "jsonlite", "ggplot2", "igraph")


# install.packages(required_packages)

# Verify
installed <- required %in% rownames(installed.packages())
print(data.frame(Package = required, Ready = installed))
```

---


## Access LLMs

Local: LM Studio

Step-by-Step Configuration

    - Download: https://lmstudio.ai (choose your OS)
    - Install a model: Search for "Mistral 7B Instruct" or "Phi-3 mini" (3-4GB VRAM)
    - Start server: Click "Local Server" → "Start Server" (port 1234)


```{r, warning=FALSE}

library(httr2)

request("http://localhost:1234/v1/models") |>
  req_perform() |>
  resp_body_json() |>
  str() 

```

---


## Minimal LM Studio Client

```{r}
library(jsonlite)

lmstudio_chat <- function(prompt) {
  body <- list(
    model = loaded_model %||% NULL,  # auto-detected by LM Studio
    messages = list(list(role = "user", content = prompt)),
    temperature = 0.3  # Lower = more deterministic
  )

  resp <- request("http://localhost:1234/v1/chat/completions") |>
    req_method("POST") |>
    req_body_json(body) |>
    req_perform()

  resp_body_json(resp)$choices[[1]]$message$content
}

```


---


## Your First LLM Query


```{r}
lmstudio_chat("Explain game theory in one sentence.")

```


---

## Structured Data from LLMs

```{r}

json <- lmstudio_chat(
  "Return 3 sorting algorithms as JSON with fields: name, stable, avg_time."
)

algos <- jsonlite::fromJSON(json)


```

---
## Wikipedia API  Dive
Fetching Clean Text

```{r}

fetch_wiki_text <- function(topic, verbose = TRUE) {
  if (verbose) cat("Fetching:", topic, "...")
  
  resp <- request("https://en.wikipedia.org/w/api.php") |>
    req_url_query(
      action = "query",
      titles = topic,
      prop = "extracts",
      exintro = TRUE,
      explaintext = TRUE,
      format = "json"
    ) |>
    req_perform()
  
  data <- resp_body_json(resp, simplifyVector = TRUE)
  pages <- data$query$pages
  page_id <- names(pages)[1]
  
  if (page_id == "-1") {
    if (verbose) cat(" NOT FOUND\n")
    return(NA_character_)
  }
  
  extract <- pages[[page_id]]$extract
  if (verbose) cat(" Done (", nchar(extract), " chars)\n", sep = "")
  
  return(extract)
}


```

---
## Output
```{r}
# Test--------------------------
algo_text <- fetch_wiki_text("Algorithm")

algo_text
```


---

## Your Turn

Fetch content for "Stack (abstract data type)" and "Queue".


---

## Create a Reusable Workflow

```{r}
# Step 1: Define topics
cs_topics <- c("Algorithm", "Data structure", "Big O notation",
               "Recursion", "Dynamic programming")

# Step 2: Fetch with progress bar
library(progress)
pb <- progress_bar$new(total = length(cs_topics))

content_list <- setNames(vector("list", length(cs_topics)), cs_topics)

for (topic in cs_topics) {
  content_list[[topic]] <- fetch_wiki_text(topic, verbose = FALSE)
  pb$tick()
}

# Step 3: Clean the data
clean_text <- function(text) {
  if (is.na(text)) return("")
  text |>
    tolower() |>
    strsplit(" ") |>
    unlist() |>
    table() |>
    sort(decreasing = TRUE) |>
    head(20)
}

# Step 4: Analyze word frequencies
word_freqs <- lapply(content_list, clean_text)

# View results
print(word_freqs$Algorithm[1:10])
```


---

## Wikipedia API Basics

```{r}


# Get raw page content
page <- "Sorting algorithm"
response <- request("https://en.wikipedia.org/w/api.php") %>%
  req_url_query(
    action = "parse",
    page = page,
    format = "json",
    prop = "text",
    formatversion = 2
  ) %>%
  req_perform() %>%
  resp_body_json(simplifyVector = TRUE)

# Extract HTML
html_content <- response$parse$text

```

---
## Parsing Wikipedia Tables

```{r}

library(rvest)

tables <- read_html(html_content) %>%
  html_table()


comparison_table <- tables[[2]]
head(comparison_table, 3)
```
---
## Batch Fetching Multiple Pages

```{r}
fetch_wiki <- function(topic) {
  Sys.sleep(0.5)  # Rate limiting
  
  tryCatch({
    # Make request
    resp <- request("https://en.wikipedia.org/w/api.php") %>%
      req_url_query(
        action = "query",
        titles = topic,
        prop = "extracts",
        exintro = TRUE,
        format = "json"
      ) %>%
      req_perform()
    
    # Parse JSON safely
    data <- resp_body_json(resp, simplifyVector = TRUE)
    
    # SAFE EXTRACTION - pages is a NAMED list like list("12345" = list(...))
    pages <- data$query$pages
    
    # Get the page ID (key) - could be "-1" if page not found
    page_id <- names(pages)[1]
    
    page_data <- pages[[page_id]]
    extract <- page_data$extract
  
    return(extract)
    
  }, error = function(e) {
    message("Wikipedia API error for '", topic, "': ", e$message)
    return(NA_character_)
  })
}


topics <- c("Algorithm", "Data structure", "embedding", 
            "ranking", "generative model")

# Loop with progress messages
content_list <- list()
for (topic in topics) {
  message("Fetching: ", topic)
  content_list[[topic]] <- fetch_wiki(topic)
  
  if (!is.na(content_list[[topic]])) {
    cat("   ✓ Got", nchar(content_list[[topic]]), "characters\n")
  } else {
    cat("   ✗ Failed\n")
  }
}

# Show results
print(content_list)
```



---
Automated Lecture Notes

```{r}

# Goal: Scrape topic → Generate code examples → Save notes
fetch_wiki <- function(topic) {
  # Rate limiting - be nice to Wikipedia
  Sys.sleep(0.5)
  
  tryCatch({
    response <- request("https://en.wikipedia.org/w/api.php") %>%
      req_url_query(
        action = "query",
        titles = topic,
        prop = "extracts",
        exintro = TRUE,
        explaintext = TRUE,  # Get plain text, no HTML
        format = "json"
      ) %>%
      req_perform() %>%
      resp_body_json(simplifyVector = TRUE)
    
    # SAFE extraction - Wikipedia returns pages as NAMED LIST
    pages <- response$query$pages
    
    # Get first page ID (pages is a list like list("12345" = list(...)))
    page_id <- names(pages)[1]
    
    # Check if page exists (ID = -1 means not found)
    if (page_id == "-1") {
      message("page not found: '", topic, "'")
      return(NA)
    }
    
    # Return extract
    extract <- pages[[page_id]]$extract
    
    if (is.null(extract)) {
      message("No extract available for: '", topic, "'")
      return(NA)
    }
    
    return(extract)
    
  }, error = function(e) {
    message("api error: ", e$message)
    return(NA)
  })
}

```
---
## Testing Lecture notes

```{r}
# TEST 
test_topic <- "Stack (abstract data type)"

step1 <- fetch_wiki(test_topic)
print(substr(step1, 1, 200))

```




---
## Security Best Practices

```{r}
# NEVER do this
unsafe_prompt <- "rm -rf /"  # LLM might suggest dangerous commands

# ALWAYS sanitize
safe_execution <- function(code) {
  # 1. Check for system calls
  if (grepl("system\\(|shell\\(", code)) {
    stop("System calls not allowed")
  }
  
  # 2. Run in sandboxed environment
  # 3. Use {safetensors} for model verification
  
  message("Code looks safe. Proceeding...")
}

# For CI/CD pipelines, use container isolation
```


---

## LLM-Powered Code Generation
Auto-Generate Examples from Text

```{r}
generate_r_example <- function(concept_name, wiki_text, chat) {
  prompt <- sprintf(
    "Based on this Wikipedia text about '%s', write a SHORT R code example.
    
    Text excerpt: %s
    
    Requirements:
    - Include comments explaining each step
    - Use base R only (no external packages)
    - Max 10 lines of code
    - Return only the code block",
    concept_name,
    substr(wiki_text, 1, 500)
  )
  
  code <- lmstudio_chat(prompt) #function style
  
  # Safety check (NEVER eval automatically)
  if (grepl("system\\(|file.remove", code)) {
    stop("Unsafe code detected!")
  }
  
  return(code)
}


```

---

## Example usage
```{r}
example <- generate_r_example("Stack", content_list$`Stack (abstract data type)`, chat)
cat(example)
```


---

## LLM-Powered Entity Extraction

```{r}
# Extract CS concepts automatically
extract_entities <- function(text, lmstudio_chat) {
  prompt <- sprintf(
    "From this CS text, extract ONLY technical terms as a comma-separated list.
    Text: %s
    
    Example output: recursion, stack overflow, Big O notation",
    substr(text, 1, 500)  # Limit length
  )
  
  entities <- lmstudio_chat(prompt)
  return(strsplit(entities, ",\\s*")[[1]])
}

# Run on algorithm summary
algorithm_entities <- extract_entities(
  content_list$Algorithm, 
  chat
)

print(head(algorithm_entities, 10))
```


---

## Network Graph of Concept Relationships

```{r}
library(tm)

# Create corpus from all Wikipedia content
corpus <- Corpus(VectorSource(content_list))

# Preprocess
dtm <- corpus %>%
  tm_map(content_transformer(tolower)) %>%
  tm_map(removeWords, stopwords("english")) %>%
  tm_map(removePunctuation) %>%
  tm_map(stripWhitespace) %>%
  DocumentTermMatrix()

# Get term frequencies
term_freq <- colSums(as.matrix(dtm))
top_terms <- sort(term_freq, decreasing = TRUE)[1:100]

```

---

## Build network
```{r}
library(igraph)
library(ggraph)
library(tidygraph)

# Build co-occurrence matrix
cooccur <- as.matrix(dtm)
cooccur <- t(cooccur) %*% cooccur
diag(cooccur) <- 0  # Remove self-loops

# Keep strong relationships only
cooccur[cooccur < 3] <- 0

# Create graph
g <- graph_from_adjacency_matrix(cooccur, mode = "undirected", weighted = TRUE)
```



---

## Plot of Entities

```{r out.width = "45%"}
g %>%
  ggraph(layout = "fr") +
  geom_edge_link(aes(size = weight), alpha = 0.2) +
  geom_node_point(size = 5) +
  geom_node_text(aes(label = name), repel = TRUE, size = 3) +
  theme_classic() +
  labs(title = "CS Concept Co-occurrence Network")


```

---

## Temporal Analysis of Page Revisions
```{r}
# evision history for activity analysis
get_revision_stats <- function(page_title) {
  rev_data <- request("https://en.wikipedia.org/w/api.php") %>%
    req_url_query(
      action = "query",
      prop = "revisions",
      titles = page_title,
      rvprop = "timestamp|user|size",
      rvlimit = 500,
      format = "json"
    ) %>%
    req_perform() %>%
    resp_body_json(simplifyVector = TRUE)
  
  revisions <- rev_data$query$pages[[1]]$revisions %>%
    mutate(timestamp = as.POSIXct(timestamp, format = "%Y-%m-%dT%H:%M:%SZ"))
  
  return(revisions)
}

# Analyze activity
algo_revisions <- get_revision_stats("Algorithm")
monthly_activity <- table(format(algo_revisions$timestamp, "%Y-%m"))


```

---

## Plot Revisions
```{r out.width = "45%"}
# activity over time
barplot(monthly_activity[1:12],
        main = "Algorithm Page: Recent Edit Activity",
        xlab = "Month", ylab = "Edits",
        col = "orange")
```

---

## Mini-Project

```{r, eval=FALSE}
# FINAL PROJECT STRUCTURE

# 1. Choose 3 topics
my_topics <- c("Hash table", "Binary tree", "Graph traversal")

# 2. Fetch content
guide <- lapply(my_topics, fetch_wiki_text) |>
  setNames(my_topics)

# 3. For each topic, generate:
#    - Summary (1 paragraph)
#    - R code example
#    - 3 related concepts
#    - Difficulty rating (1-5)

generate_study_card <- function(topic, text, chat) {
  prompt <- sprintf(
    "Create a study card for '%s' in this exact JSON format:
    {
      'topic': '%s',
      'summary': '...',
      'difficulty': 3,
      'related_topics': ['...', '...'],
      'r_example': '...'
    }
    
    Wikipedia text: %s",
    topic, topic, substr(text, 1, 800)
  )
  
  json <- chat$chat(prompt)
  jsonlite::fromJSON(json)
}


```

---


```{r, eval=FALSE}
# 4. Generate cards
study_cards <- lapply(names(guide), function(topic) {
  message("Processing: ", topic)
  generate_study_card(topic, guide[[topic]], chat)
})

# 5. Save as formatted markdown
save_study_guide <- function(cards, filename = "study_guide.md") {
  content <- c("# CS Study Guide\n\n", 
               unlist(lapply(cards, function(card) {
                 c(
                   sprintf("## %s (Difficulty: %d/5)\n", card$topic, card$difficulty),
                   sprintf("**Summary:** %s\n", card$summary),
                   sprintf("**Related:** %s\n", paste(card$related_topics, collapse = ", ")),
                   "```r", card$r_example, "```\n\n"
                 )
               })))
  
  writeLines(content, filename)
  message("Saved to: ", filename)
}

# Run it!
save_study_guide(study_cards)
```




---
## Common Issues & Fixes
```{r, eval=FALSE}
# PROBLEM 1: Connection refused
# SOLUTION: Check LM Studio is running
httr2::request("http://localhost:1234/v1/models") |>
  req_perform()

# PROBLEM 2: Empty responses
# SOLUTION: Increase timeout
chat$chat("Your prompt", timeout = 60)

# PROBLEM 3: Invalid JSON
# SOLUTION: Add retry with validation
for (i in 1:3) {
  json_str <- chat$chat("Return JSON...")
  if (jsonlite::validate(json_str)) break
  Sys.sleep(1)
}

# PROBLEM 4: Rate limits
# SOLUTION: Implement polite delays
polite_chat <- function(prompt) {
  Sys.sleep(0.5)  # Be nice to your CPU
  chat$chat(prompt)
}
```


---
## Best Practices Checklist
```{r, echo=FALSE}
practices <- data.frame(
  Do = c(
    "Cache LLM responses",
    "Validate JSON output",
    "Use polite delays",
    "Start small & test",
    "Read model docs"
  ),
  Dont = c(
    "Eval LLM code blindly",
    "Hardcode API keys",
    "Hammer Wikipedia",
    "Skip error handling",
    "Use enormous models"
  )
)

knitr::kable(practices, caption = "LLM Best Practices")
```

---

## Resources
```{r}
# Quick self-assessment
cat("During this seminar, I learned to:\n")
cat("1. Setup LM Studio: _______\n")
cat("2. Query LLMs from R: _______\n")
cat("3. Scrape Wikipedia: _______\n")
cat("4. Build a pipeline: _______\n")
cat("\nRate the difficulty: 1 2 3 4 5\n")
cat("What was unclear? _________________\n")
```

---

## Evaluation


```{r shiny-app, echo=FALSE, out.width = "45%"}
woRkshoptools::include_picture("./evaluation.png")
```





  
  
  
  
  
  
