---
title: "Data Literacy: Introduction to R"
subtitle: "LLMs as Research Assistants"
author: "Veronika Batzdorfer"
date: "2026-05-22"
presenter: Veronika
editor_options: 
  chunk_output_type: console
---
layout: true 

```{r child = "./content/config/sessions_setup.Rmd"}
```

---

class: center, middle

# LLMs as Research Assistants
## Principles, Prompts, and Verification

---

## Why Are We Talking About LLMs in an R Course?

.pull-left[
**The reality**
- LLMs are already in your workflow
- They will not replace R skills
- But only if you verify 
]

.pull-right[
**Principles**
1. Privacy first: local/open-source models
2. Trust but verify: audit every line
3. Structured grounding: Wikipedia API reduces hallucination
4. Reproducibility: prompts are part of your methods
]


---

### The Landscape: Local vs. Cloud

| Option | Setup | Cost | Privacy | Best For |
|--------|-------|------|---------|----------|
| **Ollama** (local) | Install app, pull model | Free | Perfect | Sensitive data, daily coding |
| **OpenRouter** (API) | API key, pay-per-token | Low | High (no training) | Accessing many models |
| **Hugging Face** (API) | Free tier | Free | Medium | Experimentation |
| **GitHub Copilot** | IDE plugin | Subscription | Low | Autocomplete |

**For this course:** We focus on **Ollama** (local) and **Wikipedia API** (structured knowledge) because they require zero subscriptions.

---

## Setup Check (2 Minutes)

> If you want to follow along with local models, install Ollama from ollama.com and run `ollama pull llama3.2` in your terminal. 

```{r setup-check, eval = FALSE}
# Check if you have the key packages
install.packages("ellmer")      # Posit's unified LLM interface
install.packages("httr2")       # Modern HTTP requests
install.packages("WikipediR")   # Wikipedia API wrapper
install.packages("jsonlite")    # JSON parsing
install.packages("rvest")       # HTML scraping for Wikipedia cleanup
```

```{r setup-load, message = FALSE, warning = FALSE}
library(ellmer)
library(httr2)
library(WikipediR)
library(jsonlite)
library(rvest)
library(tidyverse)
```

---

### 1: Prompt is Part of the Method

> In reproducible research, every transformation matters. If you used an LLM to generate your recoding scheme, your imputation strategy, or your model specification, that prompt is a methodological choice. Document it. A good practice: save prompts as text files in your project folder.

```{r prompt-logging, eval = FALSE}
# Save your prompt as a method file
prompt <- "You are an R data analyst. Given a dataframe with columns 
           bill_length_mm and body_mass_g, write ggplot2 code for a 
           scatter plot with species as color and a linear regression line."

writeLines(prompt, "methods/01_visualization_prompt.txt")
```

.highlight[Golden rule:] *If you cannot share your prompt, you cannot share your method.*

---

### 2: Use `ellmer` — One Interface, Many Models

**Interface:**
> The `ellmer` package is the way to talk to LLMs from R. It abstracts away provider differences. You can switch from a local Ollama model to an OpenAI API endpoint by changing one line of code. This is crucial for reproducibility — your analysis logic stays the same even if the backend changes.

.pull-left[
```{r ellmer-local, eval = FALSE}
# LOCAL: Ollama (free, runs on your laptop)
chat_local <- chat_ollama(
  model = "llama3.2",
  system_prompt = "You are a helpful R programming assistant. 
                   Return only valid R code with comments."
)
```
]

.pull-right[
```{r ellmer-cloud, eval = FALSE}
# CLOUD: OpenRouter (open-source models via API)
# chat_api <- chat_openrouter(
#   model = "meta-llama/llama-3.2-3b-instruct",
#   api_key = Sys.getenv("OPENROUTER_API_KEY")
# )
```
]

---

## Your First LLM Call from R

.pull-left[
```{r first-chat, eval = FALSE}
# Create a chat object (local Ollama)
chat <- chat_ollama(model = "llama3.2")

# Ask a focused, structured question
response <- chat$chat(
  "Write R code using ggplot2 to create a boxplot of body_mass_g 
   by species from the palmerpenguins dataset. Use theme_minimal()."
)

# The response is text — you must still run and verify it
cat(response)
```
]

.pull-right[
**Critical reminder**

The LLM returns .highlight[text], not executed code. You are still the analyst. Copy the code into a chunk, inspect it, and only then run it.

Never let an LLM execute code unsupervised in your environment.
]

---

#### 3: Structured Output — Force Precision

> LLMs are chatty. They wrap code in markdown fences, add explanations you do not need, and sometimes invent packages. `ellmer` supports structured outputs: you can force the model to return JSON with specific fields. This is essential for automated pipelines.

```{r structured-output, eval = FALSE}
# Define a JSON schema for structured responses
schema <- type_object(
  "R_code_analysis",
  code = type_string("Valid R code snippet, no markdown fences"),
  packages = type_array("Vector of package names required", type_string()),
  warnings = type_array("Vector of warnings about the code", type_string())
)

response <- chat$chat(
  "Analyze the relationship between flipper_length_mm and body_mass_g 
   in palmerpenguins. Return R code for a scatter plot with regression line.",
  type = schema
)

# Now you get a list: response$code, response$packages, response$warnings
str(response)
```

---

### 4: Grounding with Wikipedia API

> LLMs hallucinate facts. A cheap and effective way to ground them is to fetch structured knowledge from Wikipedia via the WikipediR package, then feed that text into the prompt as context. This is a primitive form of Retrieval-Augmented Generation (RAG).


---

### 4: Grounding with Wikipedia API

```{r wiki-grounding, eval = FALSE}
# Fetch Wikipedia content as grounding context
wiki_page <- page_content("en", "wikipedia", page_name = "Penguin")

# Extract the plain text introduction
wiki_text <- wiki_page$parse$text$`*`

# Build a grounded prompt
prompt <- paste0(
  "Context from Wikipedia:
", 
  substr(wiki_text, 1, 2000),  # First 2000 chars as context
  "

Task: Write a 2-sentence summary of penguin biology 
   suitable for a figure caption in a scientific report."
)

response <- chat$chat(prompt)
```

---

### Wikipedia + LLM Pipeline

.pull-left[
```{r wiki-llm-pipeline, eval = FALSE}
# Helper: clean Wikipedia HTML to plain text
get_wiki_summary <- function(topic) {
  page <- page_content("en", "wikipedia", page_name = topic)
  html <- read_html(page$parse$text$`*`)
  text <- html_text(html)
  # Return first 1500 characters of cleaned text
  substr(gsub("\s+", " ", text), 1, 1500)
}

# Fetch context
penguin_context <- get_wiki_summary("Penguin")

# Grounded generation
chat <- chat_ollama(model = "llama3.2")
caption <- chat$chat(paste(
  "Using ONLY the following context, write a 1-sentence figure caption",
  "about penguin morphological diversity for a scatter plot of",
  "bill length vs body mass. Context:", penguin_context
))

cat(caption)
```
]

.pull-right[
**Insight:** The LLM is constrained by the provided text. It cannot invent species or migration habits not in the Wikipedia article.

This is a .highlight[Retrieval-Augmented Generation (RAG)] primitive — and it is free.
]

---

### 5: Iterative Refinement — The Chat Thread


> Good analysis is iterative. `ellmer` maintains conversation state in the chat object. You can ask for a plot, then ask to modify colors, then ask to add a regression line — and the model remembers the previous code. This mirrors how you would actually work: draft, review, refine.


---
### 5: Iterative Refinement
```{r iterative-chat, eval = FALSE}
chat <- chat_ollama(model = "llama3.2")

# Turn 1: Draft
v1 <- chat$chat("Write ggplot2 code for a scatter plot of 
                bill_length_mm vs body_mass_g from palmerpenguins,
                colored by species.")
cat(v1)

# Turn 2: Refine (the chat object remembers context)
v2 <- chat$chat("Add separate linear regression lines per species 
                using geom_smooth(method='lm', se=FALSE). 
                Use a colorblind-safe palette.")
cat(v2)

# Turn 3: Polish
v3 <- chat$chat("Add proper axis labels, a title, and save the plot 
                to a 300dpi PNG using ggsave().")
cat(v3)
```

---

#### TASK 1: Prompt Engineering Challenge

##### 10 min.

**Scenario:** You need to explain Simpson's Paradox to a non-technical stakeholder using the penguins dataset.

**Your task:**
1. Open a chat with a local or API model.
2. Write a **bad prompt** (vague, no context) asking for an explanation of Simpson's Paradox.
3. Write a **good prompt** that includes:
   - The role ("You are a statistics communicator")
   - The audience ("explain to a biology undergrad")
   - The constraint ("use only the palmerpenguins dataset as example")
   - The format ("2 sentences + 1 ggplot2 code snippet")
4. Compare the two responses. Which is more useful?
5. Save both prompts to `prompts/task1_bad.txt` and `prompts/task1_good.txt`.

**Deliverable:** Paste the good response into an R chunk and verify the code runs.

---

## TASK 1: Solution Framework

.pull-left[
```{r task1-bad, eval = FALSE}
# BAD prompt
bad <- "Explain Simpson's paradox"

```
]

.pull-right[
```{r task1-good, eval = FALSE}
# GOOD prompt
good <- paste(
  "Role: You are a statistics teaching assistant.",
  "Audience: Second-year biology undergraduates who know basic R.",
  "Task: Explain Simpson's Paradox in exactly 2 sentences.",
  "Constraint: Use ONLY the palmerpenguins dataset.",
  "Format: 2 sentences + 1 self-contained ggplot2 code snippet",
  "showing how species confounds the bill_length vs body_mass relationship."
)
```
]

```{r task1-execute, eval = FALSE}
# Execute and verify
chat <- chat_ollama(model = "llama3.2")
response <- chat$chat(good)
cat(response)

# --- STUDENT VERIFICATION STEP ---
# Copy the code from the response into this chunk and run it:
# ggplot(penguins, aes(...)) + ...
```

---

### 6: LLMs for Code Explanation


> The most underrated use of LLMs is explaining code you inherited. Paste a base R snippet into the chat and ask for a tidyverse translation with line-by-line comments. This is safer than generation because the input code is ground truth.



---
### 6: LLMs for Code Explanation

.pull-left[
```{r explain-code, eval = FALSE}
# Inherited nightmare code
legacy_code <- "aggregate(body_mass_g ~ species + sex, 
                        data = penguins, 
                        FUN = function(x) c(mean = mean(x), sd = sd(x)))"

chat <- chat_ollama(model = "llama3.2")
explanation <- chat$chat(paste(
  "Translate this base R code into tidyverse (dplyr + tidyr).",
  "Explain what each original line does and why the tidyverse version is better.",
  "Code to translate:", legacy_code
))

cat(explanation)
```
]

.pull-right[
**Why explanation is safer**

- Input code is ground truth
- LLM only restructures, does not invent data
- You can verify line-by-line
- Less hallucination risk than generation from scratch
]

---

#### TASK 2: Explain & Refactor

#### 10 min.

**Scenario:** You receive this code from a collaborator. It works but is unreadable.

```{r messy-code}
messy <- "penguins[!is.na(penguins$bill_length_mm) & penguins$species=='Adelie', ]$body_mass_g"
```

**Your task:**
1. Paste the code into an LLM chat with the prompt: *"Explain this code to a beginner. Then rewrite it in tidyverse style with magrittr pipes."*
2. Run both versions and verify they return identical vectors (use `identical()`).
3. Ask the LLM: *"What happens if there are no Adelie penguins in the data? How would you make this robust?"*
4. Implement the robust version.

**Reflection:** Did the LLM catch the `filter()` vs. `[` behavior with zero rows?

---

## TASK 2: Solution

.pull-left[
```{r task2-original}
# Original (messy)
original <- penguins[!is.na(penguins$bill_length_mm) & penguins$species=='Adelie', ]$body_mass_g
```

```{r task2-refactored}
# LLM-rewriten
refactored <- penguins %>%
  filter(!is.na(bill_length_mm), species == "Adelie") %>%
  pull(body_mass_g)
```
]

.pull-right[
```{r task2-verify}
# Verify
identical(original, refactored)
```

```{r task2-robust}
# Robust version (handles zero rows gracefully)
robust <- penguins %>%
  filter(!is.na(bill_length_mm), species == "Adelie") %>%
  pull(body_mass_g)

# pull() on empty tibble returns numeric(0)
# Safer than $ which returns NULL
```
]

---

## Principle 7: When LLMs Fail — A Hall of Shame

### Failure Mode 1: Hallucinated Packages

```{r hallucination-1, eval = FALSE}
# LLM might generate:
library(ggplot3)       # Does not exist
library(palmerpenguin)  # Wrong name (should be palmerpenguins)
```

### Failure Mode 2: Hallucinated Columns

```{r hallucination-2, eval = FALSE}
# LLM might use:
ggplot(penguins, aes(x = wing_length_mm))  # Column does not exist!
# Real column is flipper_length_mm
```

---

## Failure Mode 3: Statistically Valid but Wrong

```{r hallucination-3}
# LLM-generated code that runs but is wrong:
wrong_model <- lm(body_mass_g ~ species + flipper_length_mm, 
                  data = penguins)
# This runs, but if the LLM omitted bill_length_mm which is a confounder,
# the coefficient for flipper_length_mm is biased (OVB from previous session!)
```

**The most dangerous code is code that runs.** Syntax checkers will not catch omitted variable bias. Only domain knowledge does.

---

#### TASK 3: Spot the Bug

### 10 min.

**Scenario:** An LLM generated the following "perfect" analysis pipeline. Find 3 errors or questionable choices.

**Your task:**
1. Identify 3 problems (statistical, coding, or reproducibility).
2. For each, write one sentence explaining why it is dangerous.
3. Ask a local LLM: *"Critique this code for reproducibility issues."* Does it catch the same problems?
4. Rewrite the pipeline correctly.


---
#### TASK 3: Spot the Bug

```{r buggy-llm-code, eval = FALSE}
# LLM-generated code (do not run blindly!)
library(tidyverse)

penguins_clean <- na.omit(penguins)


penguins_scaled <- penguins_clean %>%
  mutate(across(where(is.numeric), scale))


model <- lm(body_mass_g ~ .^2, data = penguins_scaled)


best_model <- step(model, direction = "both", trace = 0)

significant <- tidy(best_model) %>% filter(p.value < 0.05)
print(significant)
```

---

#### TASK 3: Solution 

| Problem | Why It Is Dangerous |
|---------|-------------------|
| `na.omit()` silently drops 2% of rows | Changes sample composition; use `drop_na()` with explicit columns instead |
| `scale()` applied to `year` | Year is categorical even if numeric; scaling it implies false continuity |
| `.^2` creates all 2-way interactions | 10 predictors → 45 interaction terms; massive overfitting on n = 333 |
| `step()` on same data | Overfitting; p-values invalid; stepwise is deprecated in modern practice |
| Filtering by p < 0.05 | Publication bias in a single model; ignores effect size and domain knowledge |


---

##### 8: LLMs + Wikipedia for Domain Exploration

.pull-left[
```{r wiki-explore, eval = FALSE}
# Step 1: Fetch related concepts from Wikipedia
topics <- c("Penguin", "Sexual_dimorphism", "Allometry")

context <- map_chr(topics, function(t) {
  page <- page_content("en", "wikipedia", page_name = t)
  html <- read_html(page$parse$text$`*`)
  text <- html_text(html)
  paste0("=== ", t, " ===\n", 
         substr(gsub("\s+", " ", text), 1, 800))
})

# Step 2: Build a synthesis prompt
prompt <- paste(
  "You are writing a biology methods section.",
  "Using ONLY the following Wikipedia excerpts,",
  "explain why we might expect body mass differences between",
  "male and female penguins, and why bill length might scale",
  "allometrically with body mass. Cite the topics used.",
  "\n\nContext:\n", paste(context, collapse = "\n\n")
)

chat <- chat_ollama(model = "llama3.2")
methods_text <- chat$chat(prompt)
cat(methods_text)
```
]

.pull-right[
**Why this works**

- Wikipedia is versioned (check page history)
- Community-edited with citations
- Free and programmatically accessible
- Constrains the LLM to existing knowledge

]

---

#### TASK 4: Build a Grounded Caption

#### 10 min.

**Scenario:** You need a figure caption for a scatter plot of `bill_depth_mm` vs `bill_length_mm` by species.

**Your task:**
1. Use `WikipediR` to fetch the article on "Beak" (or "Bird_beak").
2. Extract 500 characters of relevant text about bill morphology.
3. Write a prompt that constrains the LLM to use ONLY that text as biological context.
4. Generate a 2-sentence caption.
5. **Crucial step:** Verify that no biological claim in the caption exceeds what was in the Wikipedia text.

**Deliverable:** The caption + the Wikipedia excerpt you used, saved as a text file.

---

#### TASK 4: Solution Framework

```{r task4-solution, eval = FALSE}
# Fetch and clean
beak_page <- page_content("en", "wikipedia", page_name = "Beak")
beak_text <- html_text(read_html(beak_page$parse$text$`*`))
beak_clean <- substr(gsub("\\s+", " ", beak_text), 1, 500)

# Constrained prompt
prompt <- paste(
  "Write a 2-sentence figure caption for a scatter plot of",
  "bill_depth_mm vs bill_length_mm in penguins.",
  "You may ONLY use the following biological context.",
  "Do not add facts not present below.\n\nContext:", beak_clean
)

chat <- chat_ollama(model = "llama3.2")
caption <- chat$chat(prompt)

# Save provenance
writeLines(paste("Wikipedia source:\n", beak_clean, 
                 "\n\nGenerated caption:\n", caption),
           "output/figure_caption_provenance.txt")
```

---

### 9: Automation — Batch-Processing with LLMs


> Once you trust your prompt, you can scale it. Map over a vector of column names, dataset descriptions, or file paths, and ask the LLM to generate documentation, variable labels, or analysis code for each. This is where LLMs become genuinely productive — but only after you have validated the prompt on a single case.


---
#### 9: Automation — Batch-Processing with LLMs

```{r batch-llm, eval = FALSE}
# Auto-generate variable descriptions for a dataframe
describe_variable <- function(var_name, sample_values) {
  chat <- chat_ollama(model = "llama3.2")
  chat$chat(paste(
    "Write a 1-sentence description of this variable for a codebook.",
    "Variable name:", var_name,
    "Sample values:", paste(head(sample_values, 5), collapse = ", "),
    "Dataset: Palmer Penguins."
  ))
}

# Apply to all numeric columns
penguins_numeric <- select(penguins, where(is.numeric))
descriptions <- map_chr(names(penguins_numeric), function(v) {
  describe_variable(v, pull(penguins_numeric, v))
})

tibble(variable = names(penguins_numeric), description = descriptions)
```

**Warning:** This burns tokens/time. Use only for final documentation, not exploration.

---

#### TASK 5: The Skeptical Review

#### 15 min.

**Scenario:** A student submits this "AI-assisted" analysis for homework. Review it.

**Your review task:**
1. Identify 4 methodological errors (not syntax errors).
2. For each, state whether an LLM could plausibly have suggested it.


---


```{r homework-review, eval = FALSE}
# --- STUDENT SUBMISSION ---
# "I used ChatGPT to analyze the penguins dataset"

library(tidyverse)

# GPT said to remove outliers using IQR method
Q1 <- quantile(penguins$body_mass_g, 0.25, na.rm = TRUE)
Q3 <- quantile(penguins$body_mass_g, 0.75, na.rm = TRUE)
IQR_val <- IQR(penguins$body_mass_g, na.rm = TRUE)
penguins_no_outliers <- penguins %>%
  filter(body_mass_g >= (Q1 - 1.5*IQR_val) & body_mass_g <= (Q3 + 1.5*IQR_val))

# GPT said to normalize everything
penguins_norm <- penguins_no_outliers %>%
  mutate(across(where(is.numeric), ~ (.x - min(.x)) / (max(.x) - min(.x))))

# GPT said stepwise is best for variable selection
full <- lm(body_mass_g ~ ., data = penguins_norm)
best <- step(full, trace = 0)

# GPT said p < 0.05 means significant
results <- tidy(best) %>% filter(p.value < 0.05)
print(results)
```


---

## TASK 5: Key Points

| Issue | LLM Plausible? | Correct Approach |
|-------|---------------|------------------|
| IQR outlier removal on entire dataset | Yes (common bad advice) | Domain knowledge: penguins have real size variation by species; stratify first |
| Normalization before regression | Yes | Standardize only after train/test split; normalization destroys interpretability |
| `step()` on same data | Yes (outdated textbooks) | Theory-driven model or penalized regression (glmnet) |
| Filtering by p < 0.05 | Yes | Report full table with effect sizes|


---

### Summary

| # | Principle | Defense Against |
|---|-----------|---------------|
| 1 | Prompts are methods | Unreproducible analysis |
| 2 | One interface (`ellmer`) | Vendor lock-in |
| 3 | Structured output | Chatty, unparseable responses |
| 4 | Wikipedia grounding | Hallucinated facts |
| 5 | Iterative refinement | Premature commitment to bad code |
| 6 | Explanation > generation | Invented packages and columns |
| 7 | Skeptical review | Methodologically valid but wrong code |

---

## Further Resources

- [`ellmer` documentation](https://ellmer.tidyverse.org/) — Posit's unified LLM interface
- [Ollama](https://ollama.com/) — Run Llama, Mistral, Phi locally
- [OpenRouter](https://openrouter.ai/) — API access to open-source models
- [WikipediR](https://github.com/Ironholds/WikipediR) — R wrapper for MediaWiki API
- [Prompt Engineering Guide](https://www.promptingguide.ai/) — General principles

---

class: center, middle

# Questions?

## Remember: The LLM is your intern, not your supervisor.
